---
title: "Ideas for annotating factors: illustration using the 20 Newsgroups data set"
author: Peter Carbonetto
output: workflowr::wflow_html
---

There has been a lot of discussion about how best to "annotate" the
factors in a matrix factorization analysis. "Annotation" may mean
different things depending on the type of data, but here we will
consider a very simple aim in the context of text data: the data are
counts of words in documents, and we would like to select a small set
of words that best represents the meaning of a factor. For example, if
a factor captures "baseball", then words such as *bat*, *Yankees* and
*inning* would suggest this topic. Since we don't know in advance what
a factor represents, we would like to automate the selection of the
"most representative words", or, more generally, the "most
representative features". However, although we would like to automate
this selection process, some flexibility will be needed to accommodate
different types of data and different aims. So the annotation ideas
illustrated here have also been designed with flexibility in mind.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

## Initial setup

First, load the packages needed for this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
library(ggrepel)
```

Some additional plotting functions were developed in
[annotation_plots.R](https://github.com/stephenslab/single-cell-jamboree/blob/main/analysis/newsgroups_annotate.R)
to help us visualize the annotations:

```{r annotate-plot-functions}
source("../code/annotation_plots.R")
```

Set the seed for reproducibility.

```{r set-seed}
set.seed(1)
```

## The 20 Newsgroups data

Load the 20 Newsgroups data and the result of fitting a topic model to
these data:

```{r load-data}
load("../data/newsgroups.RData")
x <- colSums(counts > 0)
j <- which(x > 9)
counts <- counts[,j]
```

The data are counts of words in documents:

```{r inspect-data-1}
counts[1:4,1:4]
```

In total, there are over 18,000 documents and 18,000 words:

```{r inspect-data-2}
nrow(counts)
ncol(counts)
```

Only a small fraction (<1%) of the word counts are greater than zero:

```{r inspect-data-3}
nnzero(counts)/prod(dim(counts))
```

## A topic model analysis of the 20 Newsgroups data

Previously, using
[this R script](https://github.com/stephenslab/single-cell-jamboree/blob/main/scripts/compute_newsgroups_topics.R),
we fit a multinomial topic model with $K = 10$ topics to the 20
Newsgroups data set.

```{r load-results}
load("../output/newsgroups_topics.RData")
```

The "L" matrix in the topic model contains the topic proportions; that
is, each document is represented as an affine combination of 10
factors or "topics". (The proportions in each row of the L matrix add
up to 1.)

```{r memberships}
L <- poisson2multinom(pnmf)$L
head(round(L,digits = 3))
```

### Structure plots

First, let's use Structure plots (from the fastTopics package) to help
us understand the relationship between the inferred factors/topics and
the 20 Newsgroups using the existing document-level annotations.(These
document-level annotations are stored in the provided "topics"
variables, which admittedly has a confusing name.)

Most of the topics align very well with the newsgroups,

```{r structure-plot-1, fig.height=2.2, fig.width=7, results="hide", message=FALSE}
topics <- factor(topics,
  c("rec.sport.baseball","rec.sport.hockey","alt.atheism",
    "soc.religion.christian","talk.religion.misc","talk.politics.misc",
	"talk.politics.mideast","talk.politics.guns","sci.space",
	"rec.autos","rec.motorcycles","sci.electronics","sci.crypt",
    "sci.med","comp.windows.x","comp.graphics","comp.os.ms-windows.misc",
    "comp.sys.ibm.pc.hardware","comp.sys.mac.hardware","misc.forsale"))
newsgroups_topics <- paste0("k",c(1,3,6:10))
p1 <- structure_plot(L,grouping = topics,gap = 10,topics = newsgroups_topics)
p1
```

while others appear to be capturing different trends:

```{r structure-plot-2, fig.height=2.2, fig.width=7, results="hide", message=FALSE}
other_topics <- paste0("k",c(2,4,5))
p2 <- structure_plot(L,grouping = topics,gap = 10,topics = other_topics)
p2
```

See
[here](https://github.com/stephenslab/single-cell-jamboree/blob/main/output/structure_plot_newsgroups_topics.png)
and
[here](https://github.com/stephenslab/single-cell-jamboree/blob/main/output/structure_plot_other_topics.png)
for higher-resolution versions of these plots.

Note that topics 2 and 5, while appearing in many different
newsgroups, show up much more strongly in the rec.motorcycles and
talk.politics.guns newsgroups, respectively.

```{r structure-plots-ggsave, echo=FALSE}
ggsave("structure_plot_newsgroups_topics.png",p1,
       height = 2,width = 8,dpi = 200,bg = "white")
ggsave("structure_plot_other_topics.png",p2,
       height = 2,width = 8,dpi = 200,bg = "white")
```
