---
title: Initial exploration of the budding yeast data set
author: Peter Carbonetto
output: workflowr::wflow_html
---

Paper: Jackson et al eLife 2020 (doi:10.7554/eLife.51254)

Downloaded GSE125162_ALL-fastqTomat0-Counts.tsv.gz from GEO, accession
GSE125162. Added an extra column name, "id", to the header.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

First, load the packages needed for this analysis.

```{r load-pkgs}
library(Matrix)
library(tools)
library(readr)
```

Load the count data:

```{r load-data, message=FALSE}
counts <- read_tsv("../data/GSE125162_ALL-fastqTomat0-Counts.tsv.gz",
                   col_names = TRUE)
counts <- as.data.frame(counts)
ids    <- counts$id
counts <- counts[-1]
sample_info <- cbind(data.frame(id = ids),counts[6830:6834])
counts <- counts[1:6829]
counts <- as.matrix(counts)
counts <- as(counts,"CsparseMatrix")
rownames(counts) <- ids
sample_info <- transform(sample_info,
                         Genotype       = factor(Genotype),
				      	 Genotype_Group = factor(Genotype_Group),
					     Replicate      = factor(Replicate),
					     Condition      = factor(Condition))
```

Many cells have much smaller sequencing depth; out of caution, I
filter these out:

```{r filter-cells, fig.height=2.5, fig.width=3}
par(mar = c(4,4,1,1))
x <- rowSums(counts)
hist(log10(x),n = 64,xlab = "",ylab = "",main = "")
i <- which(x >= 1000)
sample_info <- sample_info[i,]
counts <- counts[i,]
```

Further, a handful of genes are expressed in only a very small number
of cells. Here I filter out genes that are expressed in fewer than 10
cells:

```{r filter-genes, fig.height=3, fig.width=3}
par(mar = c(4,4,1,1))
x <- colSums(counts > 0)
hist(log10(x),n = 64,xlab = "",ylab = "",main = "")
j <- which(x >= 10)
counts <- counts[,j]
```

Number of cells, number of samples, proportion of counts that are nonzero:

```{r examine-data-1}
nrow(counts)
ncol(counts)
mean(counts > 0)
```

Number of cells by condition:

```{r examine-data-2}
table(sample_info$Condition)
```

Finally, save the data and t-SNE results to an .Rdata file for more
convenient analysis in R:

```{r save-data, eval=FALSE}
save(list = c("sample_info","counts"),file = "yeast.RData")
resaveRdaFiles("yeast.RData")
```

TO DO:

+ Create t-SNE or UMAP plot.

+ How to get info on genes?

+ Save the data to an .Rdata file.



