
---
title: "Enhanced Visualization and Exploration of Pancreas Data using Seurat"
author: "Sagnik Nandy"
output: workflowr::wflow_html
---

In this analysis, we aim to generate an improved visualization of
pancreas data from the 2022 benchmarking study by
[Luecken et al.][luecken-2022] using the Seurat package in R. This
analysis uses the pre-processed dataset, `pancreas.RData`.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE, results = "hold", 
                      fig.align = "center", dpi = 120, message = FALSE)
```

## 1. Load Libraries

Load the necessary libraries for data analysis, plotting, and Seurat
analysis.

```{r libraries, message=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(cowplot)
library(RColorBrewer)
library(Biobase)
library(clusterSim)
library(fpc)
library(ggpubr)
library(gridExtra)
```

## 2. Define Custom Functions

Define any custom functions for the analysis. This includes a
customized plotting function for dimensional reduction (e.g., UMAP or
PCA) and a function to create an elbow plot of the top variable
features in a Seurat object.

```{r custom-functions}
# Customized dimensional reduction plot function (e.g., UMAP or PCA)
DimPlotSagnik <- function(
    data,
    dims = c(1, 2),         # Dimensions to plot (e.g., UMAP1 and UMAP2)
    cells = NULL,           # Subset of cells to plot
    cols = NULL,            # Colors for different groups
    pt.size = NULL,         # Point size for plotting
    reduction = NULL,       # Dimensional reduction technique (e.g., "umap" or "pca")
    group.by = NULL,        # Metadata column to color by
    split.by = NULL,        # Metadata column to facet by
    shape.by = NULL,        # Metadata column to shape by
    order = NULL,           # Custom order for plotting
    shuffle = FALSE,        # Option to shuffle data points to randomize plotting order
    seed = 1,               # Random seed for shuffling
    label = FALSE,          # Option to add labels to clusters
    label.size = 4,         # Size of cluster labels
    label.color = 'black',  # Color of the labels
    label.box = FALSE,      # Option to put labels in a box
    repel = FALSE,          # Use repelling labels to avoid overlap
    cells.highlight = NULL, # Subset of cells to highlight
    cols.highlight = '#DE2D26', # Color for highlighted cells
    sizes.highlight = 1,    # Size of highlighted cells
    na.value = 'grey50',    # Color for NA values
    ncol = NULL,            # Number of columns for faceting
    combine = TRUE,         # Combine plots into one
    raster = NULL,          # Option to use raster graphics
    raster.dpi = c(512, 512) # Resolution for raster graphics
) {
   if (length(x = dims) != 2) {
    stop("'dims' must be a two-length vector")
  }
  colnames(data) <- paste0("UMAP", dims)
  data <- as.data.frame(x = data)
  dims <- paste0("UMAP", dims)
  data <- cbind(data, group.by)
  orig.groups <- group.by
  group.by <- colnames(x = data)[3:ncol(x = data)]
  for (group in group.by) {
    if (!is.factor(x = data[, group])) {
      data[, group] <- factor(x = data[, group])
    }
  }
  if (!is.null(x = shape.by)) {
    data[, shape.by] <- object[[shape.by, drop = TRUE]]
  }
  if (!is.null(x = split.by)) {
    data[, split.by] <- object[[split.by, drop = TRUE]]
  }
  if (isTRUE(x = shuffle)) {
    set.seed(seed = seed)
    data <- data[sample(x = 1:nrow(x = data)), ]
  }
  plots <- lapply(
    X = group.by,
    FUN = function(x) {
      plot <- SingleDimPlot(
        data = data[, c(dims, x, split.by, shape.by)],
        dims = dims,
        col.by = x,
        cols = cols,
        pt.size = pt.size,
        shape.by = shape.by,
        order = order,
        label = FALSE,
        cells.highlight = cells.highlight,
        cols.highlight = cols.highlight,
        sizes.highlight = sizes.highlight,
        na.value = na.value,
        raster = raster,
        raster.dpi = raster.dpi
      )
      if (label) {
        plot <- LabelClusters(
          plot = plot,
          id = x,
          repel = repel,
          size = label.size,
          split.by = split.by,
          box = label.box,
          color = label.color
        )
      }
      if (!is.null(x = split.by)) {
        plot <- plot + FacetTheme() +
          facet_wrap(
            facets = vars(!!sym(x = split.by)),
            ncol = if (length(x = group.by) > 1 || is.null(x = ncol)) {
              length(x = unique(x = data[, split.by]))
            } else {
              ncol
            }
          )
      }
      plot <- if (is.null(x = orig.groups)) {
        plot + labs(title = NULL)
      } else {
        plot + labs(title = NULL)
      }
    }
  )
  if (!is.null(x = split.by)) {
    ncol <- 1
  }
  if (combine) {
    plots <- wrap_plots(plots, ncol = orig.groups %iff% ncol)
  }
  return(plots)
}

create_elbow_plot <- function(object, assay = "RNA", k = 5000) {
  # Get the high variable feature information
  hvf.info <- HVFInfo(object = object, assay = assay)
  
  # Sort the features by standardized variance in descending order
  sorted_standardized_variances <- hvf.info[order(hvf.info$variance.standardized, decreasing = TRUE), ]
  
  # Convert row names to a column named "Gene"
  sorted_standardized_variances <- sorted_standardized_variances %>%
    tibble::rownames_to_column("Gene") %>%
    mutate(Gene_Index = row_number())
  
  # Select the top k genes based on variance
  top_interest <- head(sorted_standardized_variances, k)
  
  # Create the elbow plot
  ggplot(top_interest, aes(x = Gene_Index, y = variance.standardized)) +
    geom_line() +
    labs(title = paste("Elbow Plot of Top", k, "Genes by Standardized Variance"),
         x = paste("Top", k, "Genes (ordered by variance)"),
         y = "Standardized Variance") +
    theme_minimal()
}

# Function to create an elbow plot for Seurat objects
create_elbow_plot <- function(object, assay = "RNA", k = 5000) {
  hvf.info <- HVFInfo(object = object, assay = assay)
  sorted_standardized_variances <- hvf.info[order(hvf.info$variance.standardized, decreasing = TRUE), ]
  sorted_standardized_variances <- sorted_standardized_variances %>%
    tibble::rownames_to_column("Gene") %>%
    mutate(Gene_Index = row_number())
  top_interest <- head(sorted_standardized_variances, k)
  ggplot(top_interest, aes(x = Gene_Index, y = variance.standardized)) +
    geom_line() +
    labs(
      title = paste("Elbow Plot of Top", k, "Genes by Standardized Variance"),
      x = paste("Top", k, "Genes (ordered by variance)"),
      y = "Standardized Variance"
    ) +
    theme_minimal()
}
```

## 3. Set Working Directory and Load Data

Set the working directory to the location of the `pancreas.RData`
file, and load the dataset, which contains "counts" and "sample_info".

```{r load-data}
# Load the pancreas data from the data directory
load("../data/pancreas.RData")
```

## 4. Study Size Factors

Visualize the distribution of total counts (size factors) per cell in
log scale.

```{r size-factors}
s <- rowSums(counts)
pdat <- data.frame(log_size_factor = log10(s))
ggplot(pdat, aes(log_size_factor)) +
  geom_histogram(bins = 64, col = "black", fill = "black") +
  labs(x = "Log(Size Factor)") +
  theme_cowplot(font_size = 10)
```

## 5. Study Gene Expression Levels

Examine the distribution of gene expression levels in the dataset.

```{r gene-expression}
p <- counts[counts > 0]  # Filter non-zero counts
pdat <- data.frame(log_rel_expression_level = log10(p))
ggplot(pdat, aes(log_rel_expression_level)) +
  geom_histogram(bins = 64, col = "black", fill = "black") +
  labs(x = "Log-Expression Level (Relative)") +
  theme_cowplot(font_size = 10)
```

## 6. Create Seurat Object

Create a Seurat object from the counts data, using `sample_info` as
metadata. Normalize the data with a custom scale factor based on the
mean size factor.

```{r create-seurat-object}
pancreas <- CreateSeuratObject(counts = t(counts), project = "pancreas", meta.data = sample_info)
scale_factor <- mean(s)
pancreas <- NormalizeData(pancreas, normalization.method = "LogNormalize", scale.factor = scale_factor)
```

## 7. Identify Variable Features

Identify the top 5,000 variable genes.

```{r find-variable-features}
pancreas <- FindVariableFeatures(pancreas, selection.method = "vst", nfeatures = 5000)
VariableFeaturePlot(pancreas)
```

## 8. Elbow Plot of Variable Features

Generate an elbow plot to analyze variance across the top variable features.

```{r elbow-plot}
create_elbow_plot(object = pancreas, assay = "RNA", k = 5000)
```

## 9. Select Top 1,000 Variable Features

Select the top 1,000 variable genes for further analysis.

```{r select-features}
desired_genes <- head(VariableFeatures(pancreas), 1000)
```

## 10. Scale Data and Run PCA

Scale the data for the selected features and run PCA.

```{r scale-data-pca}
pancreas <- ScaleData(pancreas, features = desired_genes)
pancreas <- RunPCA(pancreas, assay = "RNA", features = desired_genes)
ElbowPlot(pancreas, ndims = 50)
```

## 11. Run UMAP

Perform UMAP using the top 25 principal components.

```{r run-umap}
pancreas <- RunUMAP(pancreas, dims = 1:25)
umap_embeddings <- Embeddings(object = pancreas, reduction = "umap")
```

## 12. Define Colors for Cell Types

Define a color vector for distinguishing 14 different cell types.

```{r color-vector}
col_vector <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
  "#FFFF33", "#A65628", "#999999", "#66C2A5", "#FC8D62",
  "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F"
)
```

## 13. Create UMAP Plots by Cell Type and Technology

Generate UMAP plots where cell types are represented by colors and
technology by shapes.

```{r umap-plot, fig.width = 12}
# Create UMAP Plots by Cell Type and Technology
p1 <- DimPlotSagnik(umap_embeddings, group.by = sample_info$celltype, pt.size = 0.04, cols = col_vector) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("UMAP of Pancreas Cells by Cell Type")
p2 <- DimPlotSagnik(umap_embeddings, group.by = sample_info$tech, pt.size = 0.04, cols = col_vector) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("UMAP of Pancreas Cells by Technology")

# Combine plots side-by-side
combined_plot <- p1 + p2 + plot_layout(ncol = 2)
combined_plot
ggsave(
  filename = "combined_umap_plot.png",    # File name and format
  plot = combined_plot,                   # The plot object to save
  width = 15,                             # Width in inches
  height = 6,                             # Height in inches
  dpi = 300                               # Resolution in dots per inch (for high-quality output)
)
```

[luecken-2022]: https://doi.org/10.1038/s41592-021-01336-8
