
---
title: "Enhanced Visualization and Exploration of Pancreas Data using Seurat"
author: "Sagnik Nandy"
output: workflowr::wflow_html
---

In this analysis, we aim to generate an improved visualization of
pancreas data from the 2022 benchmarking study by
[Luecken et al.][luecken-2022] using the Seurat package in R. This
analysis uses the pre-processed dataset, `pancreas.RData`.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE, results = "hold", 
                      fig.align = "center", dpi = 120, message = FALSE)
```

## 1. Load Libraries

Load the necessary libraries for data analysis, plotting, and Seurat
analysis.

```{r libraries, message=FALSE}
library(Seurat)
library(tidyverse)
library(patchwork)
library(cowplot)
library(RColorBrewer)
library(Biobase)
library(clusterSim)
library(fpc)
library(ggpubr)
library(gridExtra)
```

## 2. Define Custom Functions

Define any custom functions for the analysis. This includes a
customized plotting function for dimensional reduction (e.g., UMAP or
PCA) and a function to create an elbow plot of the top variable
features in a Seurat object. I have already defined them and stored them in `Scripts` folder with the name `Customized_Plots.R`.

```{r custom-functions}
source("../Scripts/Customized_Plots.R")
```

## 3. Set Working Directory and Load Data

Load the dataset, which contains "counts" and "sample_info".

```{r load-data}
# Load the pancreas data from the data directory
load("../data/pancreas.RData")
```

## 4. Study Size Factors

Visualize the distribution of total counts (size factors) per cell in
log scale.

```{r size-factors, fig.width=4, fig.height=4}
s <- rowSums(counts)
pdat <- data.frame(log_size_factor = log10(s))
ggplot(pdat, aes(log_size_factor)) +
  geom_histogram(bins = 64, col = "black", fill = "black") +
  labs(x = "Log(Size Factor)") +
  theme_cowplot(font_size = 10)
```

## 5. Study Gene Expression Levels

Examine the distribution of gene expression levels in the dataset.

```{r gene-expression, fig.width=4, fig.height=4}
p <- counts[counts > 0]  # Filter non-zero counts
pdat <- data.frame(log_rel_expression_level = log10(p))
ggplot(pdat, aes(log_rel_expression_level)) +
  geom_histogram(bins = 64, col = "black", fill = "black") +
  labs(x = "Log-Expression Level (Relative)") +
  theme_cowplot(font_size = 10)
```

## 6. Create Seurat Object

Create a Seurat object from the counts data, using `sample_info` as
metadata. Normalize the data with a custom scale factor based on the
mean size factor.

```{r create-seurat-object}
pancreas <- CreateSeuratObject(counts = t(counts), project = "pancreas", meta.data = sample_info)
scale_factor <- mean(s)
pancreas <- NormalizeData(pancreas, normalization.method = "LogNormalize", scale.factor = scale_factor)
```

## 7. Identify Variable Features

Identify the top 5,000 variable genes.

```{r find-variable-features}
pancreas <- FindVariableFeatures(pancreas, selection.method = "vst", nfeatures = 5000)
VariableFeaturePlot(pancreas)
```

## 8. Elbow Plot of Variable Features

Generate an elbow plot to analyze variance across the top variable features.

```{r elbow-plot, fig.width=4, fig.height=4}
create_elbow_plot(object = pancreas, assay = "RNA", k = 5000)
```

## 9. Select Top 1,000 Variable Features

Select the top 1,000 variable genes for further analysis.

```{r select-features}
desired_genes <- head(VariableFeatures(pancreas), 1000)
```

## 10. Scale Data and Run PCA

Scale the data for the selected features and run PCA.

```{r scale-data-pca, fig.width=4, fig.height=4}
pancreas <- ScaleData(pancreas, features = desired_genes)
pancreas <- RunPCA(pancreas, assay = "RNA", features = desired_genes)
ElbowPlot(pancreas, ndims = 50)
```

## 11. Run UMAP

Perform UMAP using the top 25 principal components.

```{r run-umap}
pancreas <- RunUMAP(pancreas, dims = 1:25)
umap_embeddings <- Embeddings(object = pancreas, reduction = "umap")
```

## 12. Define Colors for Cell Types

Define a color vector for distinguishing 14 different cell types.

```{r color-vector}
col_vector <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
  "#FFFF33", "#A65628", "#999999", "#66C2A5", "#FC8D62",
  "#8DA0CB", "#E78AC3", "#A6D854", "#FFD92F"
)
```

## 13. Create UMAP Plots by Cell Type and Technology

Generate UMAP plots where cell types are represented by colors and
technology by shapes.

```{r umap-plot, fig.width=12, fig.height=5}
# Create UMAP Plots by Cell Type and Technology
p1 <- DimPlotSagnik(umap_embeddings, group.by = sample_info$celltype, pt.size = 0.04, cols = col_vector) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("UMAP of Pancreas Cells by Cell Type")
p2 <- DimPlotSagnik(umap_embeddings, group.by = sample_info$tech, pt.size = 0.04, cols = col_vector) +
  theme(plot.title = element_text(hjust = 0.5)) +
  ggtitle("UMAP of Pancreas Cells by Technology")

# Combine plots side-by-side
combined_plot <- p1 + p2 + plot_layout(ncol = 2)
combined_plot
ggsave(
  filename = "combined_umap_plot.png",    # File name and format
  plot = combined_plot,                   # The plot object to save
  width = 15,                             # Width in inches
  height = 6,                             # Height in inches
  dpi = 300                               # Resolution in dots per inch (for high-quality output)
)
```

[luecken-2022]: https://doi.org/10.1038/s41592-021-01336-8
