---
title: NMF analysis of the combined tissue + organoid GCA data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we explore whether non-negative matrix factorization (NMF)
identifies factors (i.e., gene expression programs) common to both the
tissues and organoids.

These results were generated using the `fit_gca_nmf.R` script, found
[here](https://github.com/stephenslab/single-cell-jamboree/blob/main/scripts/fit_gca_nmf.R).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE, warning=FALSE}
library(ggplot2)
library(cowplot)
library(fastTopics)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the results from running fitting an NMF model using flashier,
with $k = 30$:

```{r load-results}
load("../output/gca_nmf_k=30.RData")
L <- fl_nmf_ldf$L
colnames(L) <- paste0("k",1:30)
```

