---
title: NMF analysis of the combined tissue + organoid GCA data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we explore whether non-negative matrix factorization (NMF)
identifies factors (i.e., gene expression programs) common to both the
tissues and organoids.

These results were generated using the `fit_gca_nmf.R` script, found
[here](https://github.com/stephenslab/single-cell-jamboree/blob/main/scripts/fit_gca_nmf.R).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE, warning=FALSE}
library(ggplot2)
library(cowplot)
library(fastTopics)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the results from running fitting an NMF model using flashier,
with $k = 30$:

```{r load-results}
load("../output/gca_nmf_k=20.RData")
L <- fl_nmf_ldf$L
colnames(L) <- paste0("k",1:20)
```

*Add text here.*

```{r structure-plots-shared, fig.height=5, fig.width=7.5, results="hide"}
topics_shared <- c(2,5,7,9:11,16,19)
rows1 <- which(pdat$origin == "Tissue")
rows2 <- which(pdat$origin == "Org")
p1 <- structure_plot(L[rows1,topics_shared],gap = 10,
                     grouping = factor(sample_info[rows1,"celltype"])) +
  labs(y = "membership")
p2 <- structure_plot(L[rows2,topics_shared],gap = 10,
                     grouping = factor(sample_info[rows2,"celltype"])) +
  labs(y = "membership")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```
