---
title: NMF analysis of the combined tissue + organoid GCA data
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we explore whether non-negative matrix factorization (NMF)
identifies factors (i.e., gene programs) common to both the tissues
and organoids. These results were generated using the `fit_gca_nmf.R`
script, found
[here](https://github.com/stephenslab/single-cell-jamboree/blob/main/scripts/fit_gca_nmf.R).

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE, warning=FALSE}
library(ggplot2)
library(cowplot)
library(fastTopics)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the results from running fitting an NMF model using flashier,
with $k = 30$:

```{r load-results}
load("../output/gca_nmf_k=20.RData")
L <- fl_nmf_ldf$L
colnames(L) <- paste0("k",1:20)
```

For better Structure Plots, subsample the EC cells:

```{r subsample-ec}
rows1 <- which(sample_info$celltype == "EC")
rows2 <- which(sample_info$celltype != "EC")
rows1 <- sample(rows1,8000)
rows  <- c(rows1,rows2)
rows  <- sort(rows)
L     <- L[rows,]
sample_info <- sample_info[rows,]
```

These are the factors (gene programs) that are mostly specific to
tissues. Observe that the memberships in the organoids (bottom) are
mostly very small.

```{r structure-plots-tissues, fig.height=5, fig.width=7.5, results="hide", message=FALSE}
topics_tissue <- c("k2","k3","k9","k12","k14","k15","k20")
topic_colors <-
  c("#FFB300","#803E75","#FF6800","#A6BDD7","#C10020","#CEA262",
    "#93AA00","#007D34","#F6768E","#00538A","#FF7A5C","darkblue",
    "#FF8E00","yellow","cyan","#F4C800","#817066","#593315",
    "#F13A13","forestgreen")
rows1 <- which(sample_info$origin == "Tissue")
rows2 <- which(sample_info$origin == "Org")
p1 <- structure_plot(L[rows1,],gap = 10,topics = topics_tissue,
                     grouping = factor(sample_info[rows1,"celltype"]),
					 colors = topic_colors) +
  labs(y = "membership")
p2 <- structure_plot(L[rows2,],gap = 10,topics = topics_tissue,
                     grouping = factor(sample_info[rows2,"celltype"]),
					 colors = topic_colors) +
  labs(y = "membership") +
  ylim(0,1)
plot_grid(p1,p2,nrow = 2,ncol = 1)
```

*Note:* that I'm using the cell-type labels in the "celltype" column
of the meta data as a reference point for understanding the gene
programs. 

Next, these are the gene programs mostly specific to organoids.
Again, notice that the memberships are mostly very small in the
tissues:

```{r structure-plots-organoids, fig.height=5, fig.width=7.5, results="hide", message=FALSE}
topics_organoid <- c("k4","k6","k8","k11","k13","k17","k18")
rows1 <- which(sample_info$origin == "Tissue")
rows2 <- which(sample_info$origin == "Org")
p3 <- structure_plot(L[rows1,],gap = 10,topics = topics_organoid,
                     grouping = factor(sample_info[rows1,"celltype"]),
					 colors = topic_colors) +
  labs(y = "membership") +
  ylim(0,2)
p4 <- structure_plot(L[rows2,],gap = 10,topics = topics_organoid,
                     grouping = factor(sample_info[rows2,"celltype"]),
					 colors = topic_colors) +
  labs(y = "membership")
plot_grid(p3,p4,nrow = 2,ncol = 1)
```

Finally, these are the gene programs that appear in both the tissues
and the organoids. Some of the 

```{r structure-plots-shared, fig.height=4.75, fig.width=7.5, results="hide", message=FALSE}
topics_shared <- c("k5","k7","k10","k16","k19")
rows1 <- which(sample_info$origin == "Tissue")
rows2 <- which(sample_info$origin == "Org")
p5 <- structure_plot(L[rows1,],gap = 10,topics = topics_shared,
                     grouping = factor(sample_info[rows1,"celltype"]),
					 colors = topic_colors) +
  labs(y = "membership")
p6 <- structure_plot(L[rows2,],gap = 10,topics = topics_shared,
                     grouping = factor(sample_info[rows2,"celltype"]),
					 colors = topic_colors) +
  labs(y = "membership")
plot_grid(p5,p6,nrow = 2,ncol = 1)
```
