---
title: Topic modeling analysis of the GTEx v11 data set
author: Peter Carbonetto
output: workflowr::wflow_html
---

This analysis is a brief examination of the topic modeling results
generated by the `compute_gtex_factors.R` script.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages needed for this analysis:

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the GTEx meta data:

```{r load-data}
load("../data/gtex_meta_data_v11.RData")
```

Load the results of the topic modeling analysis with $K = 31$ topics:

```{r load-results}
load("../output/gtex_tm_k=31.RData")
```

A quick inspection of the topic proportions grouped by tissue shows
that indeed much of the tissue-related structure in expression is
captured by the topics. This includes closely related tissues being
captured by the same topic or combinations of the same topics (e.g.,
artery, various brain regions).

```{r structure-plot, fig.height=4.75, fig.width=8, results="hide", message=FALSE}
k <- ncol(tm$L)
topics_to_plot <- 1:k
tissues <- factor(sample_info$SMTSD)
levels(tissues) <- tolower(levels(tissues))
structure_plot(tm,topics = topics_to_plot,gap = 5,grouping = tissues) +
  theme(axis.text.x = element_text(angle = 90,hjust = 1,vjust = 0.5))
```
