---
title: "Initial exploration of the human brain dataset: committed OPC cells"
author: Junming Guan
output: workflowr::wflow_html
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#", collapse = TRUE, results = "hold",
                      fig.align = "center", dpi = 120)
```


### Introduction

This single-nucleus RNA seq dataset is from the paper "Transcriptomic diversity of cell types across the adult human brain" (Siletti, 2023). around 3 millions nuclei were collected from around 100 dissections from the following areas of brains of 3 donors:

-   Forebrain:

    -   cerebral cortex

    -   cerebral nuclei

    -   hypothalamus

    -   hippocampus formation

    -   thalamus complex

-   Midbrain

-   Hindbrain:

    -   pons

    -   cerebellum

    -   Myelencephalon (Medulla)

-   Spinal cord

The authors performed hierarchical graph-based clustering, grouping the cells into superclusters, clusters, and subclusters. The data can be accessed [here](https://data.humancellatlas.org/hca-bio-networks/nervous-system/atlases/brain-v1-0), with files organized by supercluster or by dissection.

This exploratory analysis focuses on committed oligodendrocyte precursor (COP) cells (see https://cellxgene.cziscience.com/e/f16f4108-7873-4035-9989-3748da1a7ff1.cxg/). There are 4,720 cells and 2,1462 genes (after QC) in the dataset.

```{r load-pkg, message=FALSE}
library(Matrix)
# library(MatrixExtra)
library(flashier)
library(fastTopics)
library(ggplot2)
library(cowplot)
library(dplyr)
library(readr)
library(knitr)
```

Load data and fit
```{r load-data, message=FALSE}
# Taken from https://github.com/stephenslab/pathways/blob/master/inst/code/read_gene_set_data.R
read_gene_info <- function (file) {
  
  # Read the data into a data frame.
  out <- suppressMessages(read_delim(file,delim = "\t",col_names = TRUE))
  class(out) <- "data.frame"
  dbXrefs    <- out$dbXrefs
  out        <- out[c("GeneID","Symbol","Synonyms","chromosome")]
  
  # Set any entries with a single hyphen to NA, and convert the
  # "chromosome" column to a factor.
  out$chromosome[out$chromosome == "-"] <- NA
  out$Synonyms[out$Synonyms == "-"]     <- NA
  dbXrefs[dbXrefs == "-"]               <- NA
  out <- transform(out,chromosome = factor(chromosome))
  
  # Extract the Ensembl ids. Note that a small number of genes map to
  # more than one Ensembl id; in those cases, we retain the first
  # Ensembl id only.
  dbXrefs <- strsplit(dbXrefs,"|",fixed = TRUE)
  out$Ensembl <- sapply(dbXrefs,function (x) {
    i <- which(substr(x,1,8) == "Ensembl:")
    if (length(i) > 0)
      return(substr(x[i[1]],9,nchar(x[i[1]])))
    else
      return(as.character(NA))
  })
  
  # For human genes, extract the HGNC (HUGO Gene Nomenclature
  # Committee) ids.
  out$HGNC <- sapply(dbXrefs,function (x) {
    i <- which(substr(x,1,10) == "HGNC:HGNC:")
    if (length(i) > 0)
      return(substr(x[i[1]],6,nchar(x[i[1]])))
    else
      return(as.character(NA))
  })
  
  # Return the processed gene data.
  return(out)
}

homo_sapien_geno_info <- read_gene_info('../data/Homo_sapiens.gene_info.gz')
data <- readRDS('../data/human_brain_COP_cells.rds')
counts <- t(data$RNA$data)

# Keep only protein-coding genes and remove those without nonzero counts
reduced_counts <- 
  counts[, colnames(counts) %in% homo_sapien_geno_info$Ensembl]
cols_to_keep <- colSums(reduced_counts != 0, na.rm = TRUE) > 0
reduced_counts <- reduced_counts[, cols_to_keep]
load('../data/human_brain_COP_cells_fit.RData')
```



```{r custom-func}
map_tissue <- function(tissue) {
  if (tissue %in% c("cerebral cortex", "cerebral nuclei", "hypothalamus", 
                    "hippocampal formation", "thalamic complex")) {
    return("forebrain")
  } else if (tissue == "midbrain") {
    return("midbrain")
  } else if (tissue %in% c("pons", "cerebellum", "myelencephalon")) {
    return("hindbrain")
  } else if (tissue == "spinal cord") {
    return("spinal cord")
  } else {
    return(NA)
  }
}

regions <- sapply(data$tissue, map_tissue)

```


### t-SNE and UMAP

The dataset includes precomputed tSNE and UMAP embeddings, allowing us to plot them directly. We can color the cells by tissue, by region, or by cluster.

#### t-SNE

```{r tsne}
# colors <- brewer.pal(length(unique(data$tissue)), "Paired")
colors <-  c('#756bb1', '#1c9099', '#d95f0e', '#edf8b1', '#dd1c77', '#636363', '#a1d99b', '#fa9fb5', '#fec44f', '#de2d26')
ggplot(Embeddings(data$tSNE) , aes(x = TSNE_1, y = TSNE_2, color = data$tissue)) +
  geom_point(alpha = 0.7) +
  labs(title = "t-SNE Plot Colored by Tissue Type", x = "t-SNE 1", y = "t-SNE 2") +
  theme_minimal() +
  scale_color_manual(values = colors)

# colors <- brewer.pal(length(unique(regions)), "Set1")
# ggplot(Embeddings(data$tSNE) , aes(x = TSNE_1, y = TSNE_2, color = regions)) +
#   geom_point(alpha = 0.7) +
#   labs(title = "t-SNE Plot Colored by Tissue Type", x = "t-SNE 1", y = "t-SNE 2") +
#   theme_minimal() +
#   scale_color_manual(values = colors)

# colors <- brewer.pal(length(unique(data$cluster_id)), "Paired")
colors <-  c('#756bb1', '#1c9099', '#d95f0e', '#edf8b1', '#dd1c77', '#636363', '#a1d99b')
ggplot(Embeddings(data$tSNE) , aes(x = TSNE_1, y = TSNE_2, color = data$cluster_id)) +
  geom_point(alpha = 0.7) +
  labs(title = "t-SNE Plot Colored by Cluster", x = "t-SNE 1", y = "t-SNE 2") +
  theme_minimal() +
  scale_color_manual(values = colors)
```
#### UMAP

```{r umap}
# colors <- brewer.pal(length(unique(data$tissue)), "Paired")
colors <-  c('#756bb1', '#1c9099', '#d95f0e', '#edf8b1', '#dd1c77', '#636363', '#a1d99b', '#fa9fb5', '#fec44f', '#de2d26')
ggplot(Embeddings(data$UMAP) , aes(x = UMAP_1, y = UMAP_2, color = data$tissue)) +
  geom_point(alpha = 0.7) +
  labs(title = "UMAP Plot Colored by Tissue Type", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal() +
  scale_color_manual(values = colors)

# ggplot(Embeddings(data$UMAP) , aes(x = UMAP_1, y = UMAP_2, color = regions)) +
#   geom_point(alpha = 0.7) +
#   labs(title = "UMAP Plot Colored by Tissue Type", x = "UMAP 1", y = "UMAP 2") +
#   theme_minimal() +
#   scale_color_manual(values = rainbow(length(unique(regions))))

# colors <- brewer.pal(length(unique(data$cluster_id)), "Paired")
colors <-  c('#756bb1', '#1c9099', '#d95f0e', '#edf8b1', '#dd1c77', '#636363', '#a1d99b')
ggplot(Embeddings(data$UMAP) , aes(x = UMAP_1, y = UMAP_2, color = data$cluster_id)) +
  geom_point(alpha = 0.7) +
  labs(title = "UMAP Plot Colored by Cluster", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal() +
  scale_color_manual(values = colors)
```

### Flashier

#### NMF


Redefine subcluster ids so that the hierarchical structure is clearer

```{r subclaster-ids}
# M <- table(data$subcluster_id, data$cluster_id)
# M
subcluster_id <- as.vector(data$subcluster_id)
subcluster_id[subcluster_id == '3033'] <- '37.1'
subcluster_id[subcluster_id == '3035'] <- '37.2'
subcluster_id[subcluster_id == '3036'] <- '37.3'
subcluster_id[subcluster_id == '3037'] <- '37.4'
subcluster_id[subcluster_id == '3038'] <- '37.5'

subcluster_id[subcluster_id == '3030'] <- '38.1'
subcluster_id[subcluster_id == '3031'] <- '38.2'
subcluster_id[subcluster_id == '3032'] <- '38.3'
subcluster_id[subcluster_id == '3034'] <- '38.4'

subcluster_id[subcluster_id == '3027'] <- '39.1'
subcluster_id[subcluster_id == '3028'] <- '39.2'
subcluster_id[subcluster_id == '3029'] <- '39.3'

subcluster_id[subcluster_id == '3014'] <- '41.1'
subcluster_id[subcluster_id == '3015'] <- '42.2'

subcluster_id[subcluster_id == '3007'] <- '42.1'
subcluster_id[subcluster_id == '3008'] <- '42.2'
subcluster_id[subcluster_id == '3009'] <- '42.3'
subcluster_id[subcluster_id == '3010'] <- '42.4'

subcluster_id[subcluster_id == '3006'] <- '43.1'

subcluster_id[subcluster_id == '3195'] <- '75.1'
subcluster_id[subcluster_id == '3196'] <- '75.2'
subcluster_id <- factor(subcluster_id)
```


```{r flashier-plot, message=FALSE, results='hide',fig.keep='all'}
flash_plot_scree(flashier_fit)


plot(flashier_fit, 
     plot_type = "heatmap",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     gap = 25)

plot(flashier_fit, 
     plot_type = "histogram",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 25)
```



```{r flashier-struc-plot, message=FALSE, results='hide',fig.keep='all'}
plot(flashier_fit, 
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

plot(flashier_fit, 
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = subcluster_id,
     bins = 60, gap = 100)
```

```{r examine-factors}
vals <- ldf(flashier_fit,type="m")
ncells <- colSums(vals$L>0.1)
ncells
```

Several factors are load on only a small number of cells. 

Removing those factors from the structure plots:
```{r flashier-struc-plot-reduced, message=FALSE, results='hide',fig.keep='all'}
plot(flashier_fit, 
     kset = which(ncells > 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

plot(flashier_fit, 
     kset = which(ncells > 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = subcluster_id,
     bins = 60, gap = 100)
```

Showing those factors in the structure plots:
```{r flashier-struc-plot-reduced-outlier, message=FALSE, results='hide',fig.keep='all'}
plot(flashier_fit, 
     kset = which(ncells <= 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

plot(flashier_fit, 
     kset = which(ncells <= 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = subcluster_id,
     bins = 60, gap = 100)
```
Top 10 driving genes for factors that are loading on a small number of cells

```{r driving-genes, message=FALSE, results='hide',fig.keep='all'}
low_factors <- which(ncells <= 5)
top10_genes <- list()
for (f in low_factors) {
  factor_name <- paste('Factor', f)
  genes <- head(sort(vals$F[, f], decreasing = T), 10)
  top10_ensembl <- names(genes)
  top10_symbol <- homo_sapien_geno_info$Symbol[match(top10_ensembl, homo_sapien_geno_info$Ensembl)]
  top10_genes[[factor_name]] <- top10_symbol
  
  factor_name <- paste('Factor', f, 'score')
  top10_genes[[factor_name]] <- as.vector(genes)
  
}
top10_genes %>% as.data.frame() %>% kable()
```



##### loadings on UMAP by factor

Factor 2, 3 and 4 are heavily loaded, whereas factor 8 and 9 are loaded on a single cell.

```{r umap-by-factor, message=FALSE, results='hide',fig.keep='all'}
umap <- Embeddings(data$UMAP)
source('../code/plot_loadings_on_umap.R')

for (f in c(2, 3, 4, 8, 9)) {
  loading <- vals$L[, f]
  p <- plot_loadings_on_umap(umap, loading, f)
  print(p)
}

```


#### Semi-NMF


```{r flashier-plot-cluster, message=FALSE, results='hide',fig.keep='all'}
flash_plot_scree(flashier_fit_semi)


plot(flashier_fit_semi, 
     plot_type = "heatmap",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     gap = 25)

plot(flashier_fit_semi, 
     plot_type = "histogram",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 25)
```


```{r flashier-seminmf-struc-plot, message=FALSE, results='hide',fig.keep='all'}
plot(flashier_fit_semi, 
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

plot(flashier_fit_semi, 
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = subcluster_id,
     bins = 60, gap = 100)
```

```{r examine-factors-semi}
vals_semi <- ldf(flashier_fit_semi, type="m")
ncells_semi <- colSums(vals_semi$L>0.1)
ncells_semi
```


Several factors are load on only a small number of cells.

Removing those factors from the structure plots:
```{r flashier-struc-plot-reduced-semi, message=FALSE, results='hide',fig.keep='all'}
plot(flashier_fit_semi, 
     kset = which(ncells_semi > 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

plot(flashier_fit_semi, 
     kset = which(ncells_semi > 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = subcluster_id,
     bins = 60, gap = 100)
```

Showing factors that are only loaded on a small number of cells:
```{r flashier-struc-plot-reduced-outlier-semi, message=FALSE, results='hide',fig.keep='all'}
plot(flashier_fit_semi, 
     kset = which(ncells_semi <= 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

plot(flashier_fit_semi, 
     kset = which(ncells_semi <= 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = subcluster_id,
     bins = 60, gap = 100)
```


Top 10 driving genes for factors that are loading on a small number of cells

```{r driving-genes-semi, message=FALSE, results='hide',fig.keep='all'}
low_factors <- which(ncells_semi <= 5)
top10_genes <- list()
for (f in low_factors) {
  factor_name <- paste('Factor', f)
  genes <- head(sort(vals_semi$F[, f], decreasing = T), 10)
  top10_ensembl <- names(genes)
  top10_symbol <- homo_sapien_geno_info$Symbol[match(top10_ensembl, homo_sapien_geno_info$Ensembl)]
  top10_genes[[factor_name]] <- top10_symbol
  
  factor_name <- paste('Factor', f, 'score')
  top10_genes[[factor_name]] <- as.vector(genes)
  
}
top10_genes %>% as.data.frame() %>% kable()
```

##### loadings on UMAP by factor

Factor 2 and 3 are heavily loaded, whereas factor 4 and 4 are loaded on a few cells.

```{r umap-by-factor-semi, message=FALSE, results='hide',fig.keep='all'}
umap <- Embeddings(data$UMAP)
source('../code/plot_loadings_on_umap.R')

for (f in c(2, 3, 4, 8)) {
  loading <- vals_semi$L[, f]
  p <- plot_loadings_on_umap(umap, loading, f)
  print(p)
}

```


#### Comparison

```{r struc-plot-compare, message=FALSE, results='hide',fig.keep='all'}

plot(flashier_fit, 
     kset = which(ncells > 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

plot(flashier_fit_semi, 
     kset = which(ncells_semi > 5),
     plot_type = "structure",
     pm_which = "loadings", 
     pm_groups = data$cluster_id,
     bins = 30, gap = 70)

# plot_grid(p1, p2, ncol = 1, align = "v", axis = "lr", rel_heights = c(1, 1))
```




### FastTopics


```{r fasttopics-plot-progress, message=FALSE, results='hide',fig.keep='all'}
plot_progress(fasttopics_fit,x = "iter",add.point.every = 10,colors = "black") +
  theme_cowplot(font_size = 10)

```

```{r fasttopics-loglik-plot, message=FALSE, results='hide',fig.keep='all'}
loglik <- loglik_multinom_topic_model(reduced_counts, fasttopics_fit)
pdat <- data.frame(loglik)
ggplot(pdat,aes(loglik)) +
  geom_histogram(bins = 64,color = "white",fill = "black",size = 0.25) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)

# subpop_colors <- c("dodgerblue","forestgreen","darkmagenta","skyblue","gold", )
subpop_colors <- c("red", "blue", "green", "purple", "orange", "pink", "cyan", "brown", "yellow", "darkgreen")

pdat <- data.frame(loglik = loglik,subpop = data$tissue)
ggplot(pdat,aes(x = loglik,fill = subpop)) +
  geom_histogram(bins = 64,color = "white",size = 0.25) +
  scale_fill_manual(values = subpop_colors) +
  labs(y = "number of cells") +
  theme_cowplot(font_size = 10)
```


```{r fasttopics-struc-plot, message=FALSE, results='hide',fig.keep='all'}
structure_plot(fasttopics_fit, grouping = data$cluster_id, gap = 70)
structure_plot(fasttopics_fit, grouping = subcluster_id, gap = 70)
```
```{r examine-factors-fasttopics}
ncells_ft <- colSums(fasttopics_fit$L>0.1)
ncells_ft
```
No factors that are loaded on a single cells. Still there are a couple that are loaded on a small number of cells, e.g., k8.


### GBCD

TODO

```{r fit-gbcd}
# source("../code/fit_cov_ebnmf.R")
# 
# fit.gbcd <-
#   flash_fit_cov_ebnmf(Y = reduced_counts, Kmax = 7,
#                       prior = flash_ebnm(prior_family = "generalized_binary",
#                                          scale = 0.04),
#                       extrapolate = FALSE)
```
