---
title: NMF analysis of the LPS data set
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we will revisit the LPS data set that we analyzed using a topic
model in the [Takahama et al Nat Immunol paper][takahama2024] (LPS =
lipopolysaccharide). I believe some interesting insights can be gained
by examining this data set more deeply.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load packages used to process the data, perform the analyses, and
create the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(readr)
library(data.table)
library(fastTopics)
library(NNLM)
library(ebnm)
library(flashier)
library(pathways)
library(singlecelljamboreeR)
library(ggplot2)
library(ggrepel)
library(cowplot)
```

## Prepare the data for analysis with fastTopics and flashier

Load the RNA-seq counts:

```{r load-counts}
read_lps_data <- function (file) {
  counts <- fread(file)
  class(counts) <- "data.frame"
  genes <- counts[,1]
  counts <- t(as.matrix(counts[,-1]))
  colnames(counts) <- genes
  samples <- rownames(counts)
  samples <- strsplit(samples,"_")
  samples <- data.frame(tissue    = sapply(samples,"[[",1),
                        timepoint = sapply(samples,"[[",2),
                        mouse     = sapply(samples,"[[",3))
  samples <- transform(samples,
                       tissue    = factor(tissue),
					   timepoint = factor(timepoint),
					   mouse     = factor(mouse))
  return(list(samples = samples,counts = counts))
}
out <- read_lps_data("../data/lps.csv.gz")
samples <- out$samples
counts  <- out$counts
rm(out)
```

Remove a sample that appears to be an outlier based on the NMF analyses:

```{r filter-outliers}
i       <- which(rownames(counts) != "iLN_d2_20")
samples <- samples[i,]
counts  <- counts[i,]
```

Remove genes that are expressed in fewer than 5 samples:

```{r filter-genes}
j <- which(colSums(counts > 0) > 4)
counts <- counts[,j]
```

This is the dimension of the data set we will analyze:

```{r dim-counts}
dim(counts)
```

For the Gaussian-based analyses, we will need the shifted log counts:

```{r shifted-log-counts}
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
shifted_log_counts <- log1p(counts/(a*s))
```

## Topic model (fastTopics)

First let's fit a topic model with $K = 9$ topics to the counts. This
is probably an insufficient number of topics to fully capture the
interesting structure in the data, but this is done on purpose since I
want to illustrate how the topic model prioritizes the structure.

```{r fit-topic-model-k9, cache=TRUE}
set.seed(1)
tm_k9 <- fit_poisson_nmf(counts,k = 9,init.method = "random",method = "em",
                         numiter = 20,verbose = "none",
		   			     control = list(numiter=4,nc=8,extrapolate=FALSE))
tm_k9 <- fit_poisson_nmf(counts,fit0 = tm_k9,method = "scd",numiter = 40,
                         control = list(numiter = 4,nc = 8,extrapolate = TRUE),
					     verbose = "none")
```

Structure plot comparing the topics to the tissue types:

```{r structure-plot-topic-model-k9, fig.height=1.5, fig.width=7}
rows <- order(samples$timepoint)
topic_colors <- c("magenta","darkorange","darkblue","forestgreen",
                  "dodgerblue","gray","red","olivedrab","darkmagenta",
                  "sienna","limegreen","royalblue","lightskyblue",
				  "gold")
samples <- transform(samples,
  tissue = factor(tissue,c("PBMC","BM","CO","SI","iLN","SP",
                           "TH","SK","BR","LI","HE","KI","LU")))
structure_plot(tm_k9,grouping = samples$tissue,gap = 4,
               topics = 1:9,colors = topic_colors,
   		       loadings_order = rows) +
  labs(fill = "") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5))
```

Abbreviations used: BM = bone marrow; BR = brain; CO = colon; HE =
heart; iLN = inguinal lymph node; KI = kidney; LI = liver; LU = lung;
SI = small intestine; SK = skin; SP = spleen; TH = thymus.

The topics largely correspond to the different tissues, although
because there are more tissues than topics, some tissues that are more
similar to each other shared the same topic. It is also interesting
that, for the most part, none of the topics are capturing changes
downstream of the LPS treatment. So presumably these expression
changes are more subtle than the differences in expression among the
tissues.

Fit a topic model with $K = 14$ topics to the counts:

```{r fit-topic-model, cache=TRUE}
set.seed(1)
tm <- fit_poisson_nmf(counts,k = 14,init.method = "random",method = "em",
                      numiter = 20,verbose = "none",
					  control = list(numiter = 4,nc = 8,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 40,
                      control = list(numiter = 4,nc = 8,extrapolate = TRUE),
					  verbose = "none")
```

Structure plot comparing the topics to the tissue types:

```{r structure-plot-topic-model, fig.height=1.5, fig.width=7}
structure_plot(tm,grouping = samples$tissue,gap = 4,
               topics = 1:14,colors = topic_colors,
   		       loadings_order = rows) +
  labs(fill = "") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5),
        legend.key.height = unit(0.01,"cm"),
        legend.key.width = unit(0.2,"cm"),
        legend.text = element_text(size = 6))
```

This next structure plot better highlights the topics that are
capturing expression changes over time, some being presumably driven
by the LPS-induced sepsis:

```{r structure-plot-topic-model-2, fig.height=1.5, fig.width=7}
topic_colors <- c("magenta","gray50","gray65","gray40",
                  "gray85","gray75","red","gray80","gray90",
                  "gray60","limegreen","gray70","gray55",
				  "gold")
structure_plot(tm,grouping = samples$tissue,gap = 4,
               topics = 1:14,colors = topic_colors,
			   loadings_order = rows) +
  labs(fill = "") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5),
        legend.key.height = unit(0.01,"cm"),
        legend.key.width = unit(0.2,"cm"),
        legend.text = element_text(size = 6))
```

## EBNMF (flashier)

Similar to the topic modeling analysis, let's start by fitting an
EBNMF to the shifted log counts using flashier, first with $K = 9$.
Since the greedy initialization does not seem to work well in this
example, I'll use a different initialization strategy: obtain a "good"
initialization using the NNLM package, then use this initialization to
fit a NMF using flashier. This approach is implemented in the
following function:

```{r flashier-nmf-function}
flashier_nmf <- function (X, k = 3, n.threads = 1) {
  n <- nrow(X)
  m <- ncol(X)
  nmf0 <- nnmf(X,k = 1,loss = "mse",method = "scd",
               max.iter = 10,verbose = 0,n.threads = n.threads)
  W0 <- nmf0$W
  H0 <- nmf0$H
  W0 <- cbind(W0,matrix(runif(n*(k-1)),n,k-1))
  H0 <- rbind(H0,matrix(runif(m*(k-1)),k-1,m))
  nmf <- nnmf(X,k,init = list(W = W0,H = H0),loss = "mse",method = "scd",
              max.iter = 10,verbose = 0,n.threads = n.threads)
  x  <- rpois(1e7,1/n)
  s1 <- sd(log(x + 1))
  out <- flash_init(X,var_type = 2,S = s1)
  out <- flash_factors_init(out,list(nmf$W,t(nmf$H)),ebnm_point_exponential)
  out <- flash_backfit(out,extrapolate = FALSE,maxiter = 100,verbose = 0)
  return(flash_backfit(out,extrapolate = TRUE,maxiter = 100,verbose = 0))
}
```

Now fit an NMF to the shifted log counts, with $K = 9$:

```{r flashier-nmf-k9, cache=TRUE}
set.seed(1)
fl_nmf_k9 <- flashier_nmf(shifted_log_counts,k = 9,n.threads = 8)
```

Structure plot comparing the topics to the tissue types:

```{r structure-plot-flashier-nmf-k9, fig.height=2, fig.width=7}
rows <- order(samples$timepoint)
topic_colors <- c("powderblue","dodgerblue","olivedrab","limegreen",
                  "forestgreen","red","darkmagenta","gray","darkorange",
                  "cyan","royalblue","darkblue","lightskyblue",
                  "gold","sienna")
L <- ldf(fl_nmf_k9,type = "i")$L
structure_plot(L,grouping = samples$tissue,gap = 4,
               topics = 1:9,colors = topic_colors,
			   loadings_order = rows) +
  labs(fill = "",y = "membership")
```

Like the topic model, the EBNMF model with $K = 9$ does not capture
any changes downstream of the LPS-induced sepsis.

Next fit an NMF to the shifted log counts using flashier, with $K =
15$:

```{r flashier-nmf-k15, cache=TRUE, warning=FALSE}
set.seed(1)
fl_nmf <- flashier_nmf(shifted_log_counts,k = 15,n.threads = 8)
```

Structure plot comparing the factors to the tissue types:

```{r structure-plot-flashier-nmf, fig.height=2, fig.width=7}
L <- ldf(fl_nmf,type = "i")$L
structure_plot(L,grouping = samples$tissue,gap = 4,
               topics = 1:15,colors = topic_colors,
			   loadings_order = rows) +
  labs(fill = "",y = "membership") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5),
        legend.key.height = unit(0.01,"cm"),
        legend.key.width = unit(0.25,"cm"),
        legend.text = element_text(size = 7))
```

This next structure plot better highlights the topics that capture the
processes that are driven or may be driven by LPS-induced sepsis:

```{r structure-plot-flashier-nmf-2, fig.height=2, fig.width=7}
rows <- order(samples$timepoint)
topic_colors <- c("gray95","gray70","gray80","gray50",
                  "gray60","red","gray75","gray","gray85",
                  "gray90","gray65","darkblue","gray45",
                  "gray35","gray75")
L <- ldf(fl_nmf,type = "i")$L
structure_plot(L,grouping = samples$tissue,gap = 4,
                    topics = 1:15,colors = topic_colors,
					loadings_order = rows) +
  labs(fill = "",y = "membership") +
  theme(axis.text.x = element_text(angle = 0,hjust = 0.5),
        legend.key.height = unit(0.01,"cm"),
        legend.key.width = unit(0.25,"cm"),
        legend.text = element_text(size = 7))
```

## Factors isolating responses to LPS-induced sepsis

From the Structure plots, it appears that topic 7, and possibly topic
1, are capturing processes activated by LPS. However, I conjecture
that it is difficult to determine which genes should members of 
topic 1 and which are members of the colon and small intensine
topics. Indeed, topic 1 shares many genes with these two topics:

```{r compare-topics-1, fig.height=2.25, fig.width=5}
pdat <- cbind(data.frame(gene = colnames(counts)),
              poisson2multinom(tm)$F)
rows <- which(pdat$k1 < 0.01)
pdat[rows,"gene"] <- ""
p1 <- ggplot(pdat,aes(x = k1,y = k4,label = gene)) +
  geom_point() +
  geom_text_repel(color = "gray",size = 2.5,max.overlaps = Inf) +
  scale_x_continuous(trans = "sqrt") +
  scale_y_continuous(trans = "sqrt") +
  theme_cowplot(font_size = 10)
p2 <- ggplot(pdat,aes(x = k1,y = k13,label = gene)) +
  geom_point() +
  geom_text_repel(color = "gray",size = 2.5,max.overlaps = Inf) +
  scale_x_continuous(trans = "sqrt") +
  scale_y_continuous(trans = "sqrt") +
  theme_cowplot(font_size = 10)
plot_grid(p1,p2,nrow = 1,ncol = 2)
```

Still, it is interesting that three genes, *Reg3b*, *Reg3g* and
*Lypd8*, stand out in topic 1 as distinct from the colon and SI
topics. Let's now contrast this to the situation for topic 7:

```{r compare-topics-2, fig.height=2.25, fig.width=5}
pdat <- cbind(data.frame(gene = colnames(counts)),
              poisson2multinom(tm)$F)
rows <- which(pdat$k7 < 0.008)
pdat[rows,"gene"] <- ""
p1 <- ggplot(pdat,aes(x = k7,y = k6,label = gene)) +
  geom_point() +
  geom_text_repel(color = "gray",size = 2.5,max.overlaps = Inf) +
  scale_x_continuous(trans = "sqrt") +
  scale_y_continuous(trans = "sqrt") +
  theme_cowplot(font_size = 10)
p2 <- ggplot(pdat,aes(x = k7,y = k8,label = gene)) +
  geom_point() +
  geom_text_repel(color = "gray",size = 2.5,max.overlaps = Inf) +
  scale_x_continuous(trans = "sqrt") +
  scale_y_continuous(trans = "sqrt") +
  theme_cowplot(font_size = 10)
plot_grid(p1,p2,nrow = 1,ncol = 2)
```

For illustration, I compared topic 7 to the kidney and lung
topics. The key point here is that the topic model has selected genes
for topic 7 that are very independent of the tissue topics. So this
looks quite promising. Let's now see if the result is similar for the
EBNMF model fitted to the shifted log counts:

```{r sepsis-tm-vs-nmf, fig.height=2.5, fig.width=4}
F <- ldf(fl_nmf,type = "i")$F
colnames(F) <- paste0("k",1:15)
pdat <- data.frame(tm   = poisson2multinom(tm)$F[,"k7"],
                   nmf  = F[,"k6"],
				   gene = rownames(F))
rows <- which(with(pdat,tm < 0.005 & nmf < 0.8))
pdat[rows,"gene"] <- ""
ggplot(pdat,aes(x = (tm)^(1/3),y = nmf,label = gene)) +
  geom_point() +
  geom_text_repel(color = "gray",size = 2.5,max.overlaps = Inf) +
  labs(x = "topic 7",y = "factor 6") +
  theme_cowplot(font_size = 12)
```

Indeed, factor 6 and topic 7 are cpaturing very similar expression
patterns.

Next I ran GSEA on the on factor 6. (Running GSEA on topic 7 is
complicated by the fact that it would be better to "shrink" the
estimates before running GSEA, whereas this was automatically done for
EBNMF result.)

```{r gsea-fl-nmf-k6}
data(gene_sets_mouse)
gene_sets     <- gene_sets_mouse$gene_sets
gene_info     <- gene_sets_mouse$gene_info
gene_set_info <- gene_sets_mouse$gene_set_info
j <- which(with(gene_sets_mouse$gene_set_info,
                (database == "MSigDB-C2" &
                 grepl("CP",sub_category_code,fixed = TRUE)) |
                (database == "MSigDB-C5") &
                 grepl("GO",sub_category_code,fixed = TRUE)))
genes <- sort(intersect(rownames(F),gene_info$Symbol))
i     <- which(is.element(gene_info$Symbol,genes))
gene_info     <- gene_info[i,]
gene_set_info <- gene_set_info[j,]
gene_sets     <- gene_sets[i,j]
rownames(gene_sets) <- gene_info$Symbol
rownames(gene_set_info) <- gene_set_info$id
gene_set_info <- gene_set_info[,-2]
F <- ldf(fl_nmf,type = "i")$F
colnames(F) <- paste0("k",1:15)
gsea_fl_nmf <- singlecelljamboreeR::perform_gsea(F[,"k6"],gene_sets,
                                                 gene_set_info,L = 15,
                                                 verbose = FALSE)
out <- gsea_fl_nmf$selected_gene_sets
out$top_genes <- sapply(out$top_genes,function (x) paste(x,collapse = " "))
write_csv(out,"lps_gsea_fl_nmf.csv",quote = "none")
```

The top gene set is the IFN-$\alpha/\beta$ signaling pathway, but
other gene sets clearly relate to inflammation and immune system
function:

```{r  gsea-fl-nmf-k6-summary, fig.height=2.5, fig.width=4}
print(gsea_fl_nmf$selected_gene_sets[c(2:7,9)],n = Inf)
```

This is the same scatterplot as the one just above, but with the genes
in the IFN-$\alpha/\beta$ signaling pathway highlighted:

```{r sepsis-tm-vs-nmf-with-gsea, fig.height=2.5, fig.width=5}
pdat$pathway <- FALSE
pathway_genes <- names(which(gene_sets[,"M973"] > 0))
pdat[pathway_genes,"pathway"] <- TRUE
pdat <- pdat[order(pdat$pathway),]
ggplot(pdat,aes(x = (tm)^(1/3),y = nmf,label = gene,color = pathway)) +
  geom_point() +
  geom_text_repel(color = "gray",size = 2.5,max.overlaps = Inf) +
  scale_color_manual(values = c("darkblue","orangered")) +
  labs(x = "topic 7",y = "factor 6") +
  theme_cowplot(font_size = 12)
```

[takahama2024]: https://doi.org/10.1038/s41590-023-01722-8
