---
title: Exploring Neighbor joining in factorization
author: William Denault
output: workflowr::wflow_html
---

This goal of this analysis is to apply Neighbor joining to  different matrix
factorization approaches to the [pancreas data set](pancreas.html) and
assess how well it works for this data set.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

First, load the packages needed for this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
library(ape)
```

Set the seed for reproducibility.

```{r set-seed}
set.seed(1)
```

This is a function I will use below to subsample the cell types that
have a lot of cells so that the rare cell types are more visible in
the Structure plots.

```{r subsample-cell-types}
subsample_cell_types <- function (x, n = 1000) {
  cells <- NULL
  groups <- levels(x)
  for (g in groups) {
    i  <-  which(x == g)
    n0 <- min(n,length(i))
    i  <- sample(i,n0)
    cells <- c(cells,i)
  }
  return(sort(cells))
}
```

Load the pancreas data set and the outputs generated from running the
`compute_pancreas_factors.R` and `compute_pancreas_factors2.R` scripts.

```{r load-data}
load("../data/pancreas.RData")
load("../output/pancreas_factors.RData")
timings0 <- timings
load("../output/pancreas_factors2.RData")
timings <- c(timings0,timings)
```
 

### flashier NMF

Reproduction of Peter resulst

```{r structure-plots-flashier-nmf, results="hide", message=FALSE, fig.height=5.5, fig.width=8}
cells <- subsample_cell_types(sample_info$celltype,n = 500)
L <- fl_nmf_ldf$L
k <- ncol(L)
colnames(L) <- paste0("k",1:k)
batch_factors    <- c(2:5,7:8,20)
celltype_factors <- c(6,11:19,21)
other_factors    <- c(1,9:10,22:23)
celltype <- sample_info$celltype
celltype <-
 factor(celltype,
        c("acinar","ductal","activated_stellate","quiescent_stellate",
		  "endothelial","macrophage","mast","schwann","t_cell","alpha",
		  "beta","delta","gamma","epsilon"))
p1 <- structure_plot(L,topics = batch_factors,grouping = sample_info$tech,
                     gap = 10,perplexity = 70) +
  labs(y = "membership",title = "data-set factors",
       fill = "factor",color = "factor")
p2 <- structure_plot(L[cells,],topics = celltype_factors,
                     grouping = celltype[cells],gap = 25,
					 n = Inf,perplexity = 70) +
  labs(y = "membership",title = "cell-type factors",
       fill = "factor",color = "factor")
p3 <- structure_plot(L[cells,],topics = other_factors,
                     grouping = celltype[cells],gap = 25,
					 n = Inf,perplexity = 70) +
  labs(y = "membership",title = "other factors",
       fill = "factor",color = "factor")
plot_grid(p1,p2,p3,nrow = 3,ncol = 1)
```


Let's start by applying NJ to all the factor together.
It generates a couple of negative edge length that we correct


```{r NJ NMF_all}

dist_mat= t(fl_nmf_ldf$F)%*%fl_nmf_ldf$F
dimnames(dist_mat)= list(1:ncol(dist_mat), 1:ncol(dist_mat))
check_tree=nj(dist_mat )
plot(check_tree, main="NMF +NJ")
check_tree$edge.length=abs(check_tree$edge.length)
plot(check_tree)
 

```


Lets now focus just on the so called celltype_factors and apply NJ

```{r}
celltype_factors <-c(6,11:19,21)

check_tree=nj(dist_mat[celltype_factors, celltype_factors] )
plot(check_tree, main="Cell factor + NMF +NJ")
check_tree$edge.length=abs(check_tree$edge.length)
check_tree
plot(check_tree)

```
In this case there are no negative edge 

Note that factors that are close on the tree tend not to appear together
in the "factor plot" above (eg. 11 and 6 or 14 and 15)

In other words if a cell is loaded in one of the branch it tends not to be
loaded in a branch within its vicinity.  

 
### Topic model (fastTopics)
 Peter script

```{r structure-plots-fasttopics, results="hide", message=FALSE, fig.height=5.5, fig.width=8}
L <- poisson2multinom(pnmf)$L
batch_topics     <- c(2,5,6,11,12)
celltype_topics  <- c(4,9,8,15,16,17,18,19,20)
celltype_topics2 <- c(1,3,7,10,13,14,21,22,23)
p1 <- structure_plot(L[,batch_topics],grouping = sample_info[,"tech"],
                     gap = 10,perplexity = 70) +
  ggtitle("data-set topics")
p2 <- structure_plot(L[cells,celltype_topics],grouping = celltype[cells],
                     gap = 25,perplexity = 70,n = Inf) +
  ggtitle("cell-type topics")					 
p3 <- structure_plot(L[cells,celltype_topics2],grouping = celltype[cells],
                     gap = 25,perplexity = 70,n = Inf) +
  ggtitle("more cell-type topics")					 
plot_grid(p1,p2,p3,nrow = 3,ncol = 1)
```

 
Same let's run NJ on all the factor and then let's focus of the cell topic 
 
 
 
```{r}
dist_mat=  t(pnmf$F)%*% pnmf$F
dimnames(dist_mat)= list(1:ncol(dist_mat), 1:ncol(dist_mat))
check_tree=nj(dist_mat   )
 
plot(check_tree, main="Topic +NJ")
check_tree$edge.length=abs(check_tree$edge.length)
check_tree
plot(check_tree)






```
 
Clearly the tree do not make sens.
Let's look at the celltopic only 


 
```{r}

check_tree=nj( dist_mat[celltype_topics, celltype_topics]    )
 
plot(check_tree, main="Cell topic + FastTopic +NJ")
check_tree$edge.length=abs(check_tree$edge.length)
check_tree
plot(check_tree)

```
 
 
Unfortunately this tree still has some problem.
However we notice the same pattern as before.
What is close on the tree tend not to co-occur on the "factor plot"
In other words if a cell is loaded in one of the branch it tend not to be
loaded in a branch within its vicinity. (eg 4 and 9  or  18 and 19)
 
  
