---
title: "pancreas_celseq2_ica_02"
author: "Matthew Stephens"
date: "2026-09-16"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    toc_depth: 2
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

A second look at ICA on the pancreas CEL-seq2 data, incorporating lessons from the first analysis (`pancreas_celseq2_ica.Rmd`). Key differences from the first analysis:

- We focus on contrast functions chosen to favour one-sided/sparse sources, and avoid combining groups. For example, 
$G(x) = x^3$ (skewness), $G(x) = \log\Phi(\alpha x)$ (log-normal CDF), and $G(x) = x|x|$ (the tilting term of TLC without the log-cosh).
This means we can center the data without breaking things, so we no longer need to augment by an intercept.

- We focus on **rank-$r$ fastICA**, whitening to 30 PCs and keeping all $r = 30$ components.

- We use both polar and Gram Schmidt to do orthogonalization.
In GS each iteration applies the contrast update followed by Gram-Schmidt reorthogonalization. Components are ordered by decreasing objective before GS, so the best-objective direction is processed first and kept exactly; subsequent components are deflated against all preceding ones. After GS the columns remain orthonormal.

```{r load-data}
library(Matrix)
library(ggplot2)

load("../data/pancreas.RData")
set.seed(1)

i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i, ]
counts      <- counts[i, ]

x      <- colSums(counts > 0)
j      <- which(x > 9)
counts <- counts[, j]

a <- 1
s <- rowSums(counts)
s <- s / mean(s)
Y <- MatrixExtra::mapSparse(counts / (a * s), log1p)
Y <- scale(Y, scale = FALSE)   # centre columns
```

## Preprocessing: SVD of centred data

```{r svd}
n      <- nrow(Y)
n.comp <- 30   # PCs retained for whitening
r      <- 30   # components extracted by ICA (r <= n.comp)

Y.svd <- svd(Y, nu = n.comp, nv = 0)
# U: n x n.comp, orthonormal columns (left singular vectors of centred Y)
U_c <- Y.svd$u[, 1:n.comp]
# Whitened row space (n.comp x n, as used by rank-r fastICA)
U_w <- sqrt(n) * t(U_c)
```

Here `r r` components are extracted from a `r n.comp`-dimensional whitened
space, so each `W` is square (`r n.comp` $\times$ `r r`) with orthonormal
columns and the components span the whole whitened space. Setting `r` below
`n.comp` would instead let ICA keep only the most interesting directions and
leave the remainder unmodelled.

## Helper functions

```{r helpers}
celltype_palette <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
  "#A65628", "#F781BF", "#1B9E77", "#D95F02", "#7570B3",
  "#E7298A", "#66A61E", "#E6AB02", "#A6761D", "#666666"
)

# Polar factor via eigendecomposition (avoids SVD column reordering).
polar <- function(W) {
  eig      <- eigen(t(W) %*% W, symmetric = TRUE)
  Ainvhalf <- eig$vectors %*%
               diag(1 / sqrt(pmax(eig$values, 1e-14))) %*%
               t(eig$vectors)
  W %*% Ainvhalf
}
```

## Contrast functions and the fastICA driver

A contrast is specified by the function $G$ whose column mean the update
maximizes. `deriv()` returns $G'$ and $G''$ evaluated at $P = U^\top W$ (the
`G` and `G2` terms of the symmetric fastICA update) and `obj()` returns
$\mathrm{colMeans}(G(L))$.

```{r contrast-defs}
contrast_defs <- list(
  # G(x) = x^3:  G' = 3x^2,  G'' = 6x
  cubic = list(
    deriv = function(P) list(G = 3 * P^2, G2 = 6 * P),
    obj   = function(L) colMeans(L^3)
  ),
  # G(x) = log Phi(alpha*x),  h = phi/Phi
  logphi = list(
    deriv = function(P, alpha = 2) {
      u <- alpha * P
      h <- exp(dnorm(u, log = TRUE) - pnorm(u, log.p = TRUE))
      list(G = alpha * h, G2 = alpha^2 * (-u * h - h^2))
    },
    obj = function(L, alpha = 2) colMeans(pnorm(alpha * L, log.p = TRUE))
  ),
  # G(x) = x|x|:  G' = 2|x|,  G'' = 2 sign(x)
  skew = list(
    deriv = function(P) list(G = 2 * abs(P), G2 = 2 * sign(P)),
    obj   = function(L) colMeans(L * abs(L))
  )
)

# Gram-Schmidt reorthogonalization ordered by decreasing objective.
# W: n.comp x r with orthonormal columns; L: n x r loadings (= t(U) %*% W).
# The best-objective direction is processed first and so is kept exactly;
# each later column is deflated against all preceding ones.
gs_reorth_W <- function(W, L, obj_fn) {
  ord <- order(obj_fn(L), decreasing = TRUE)
  V   <- W[, ord, drop = FALSE]
  for (j in seq_len(ncol(V))) {
    if (j > 1)
      for (l in seq_len(j - 1))
        V[, j] <- V[, j] - sum(V[, j] * V[, l]) * V[, l]
    nrm <- sqrt(sum(V[, j]^2))
    if (nrm > 1e-10) V[, j] <- V[, j] / nrm
  }
  V
}

# Rank-r symmetric fastICA. contrast picks an entry of contrast_defs; orth is
# the orthogonalization applied after each update, either the polar factor or
# Gram-Schmidt ordered by objective. Returns the n x r loadings t(U) %*% W.
run_fastica <- function(U, contrast, orth = c("polar", "gs"),
                        r = nrow(U), n_iter = 50, seed = 1) {
  orth <- match.arg(orth)
  cf   <- contrast_defs[[contrast]]
  set.seed(seed)
  W <- polar(matrix(rnorm(nrow(U) * r), nrow(U), r))
  for (i in seq_len(n_iter)) {
    d <- cf$deriv(t(U) %*% W)
    W <- U %*% d$G - sweep(W, 2, colSums(d$G2), "*")
    W <- if (orth == "polar") polar(W)
         else gs_reorth_W(W, t(U) %*% W, cf$obj)
  }
  t(U) %*% W
}
```

## Running all six methods

Every method below is the same driver with a different `contrast` / `orth`
pair, all from the same random start (`seed = 1`) and the same number of
iterations.

```{r run-all}
method_spec <- list(
  "cubic"     = list(contrast = "cubic",  orth = "polar"),
  "logphi"    = list(contrast = "logphi", orth = "polar"),
  "skew"      = list(contrast = "skew",   orth = "polar"),
  "cubic-GS"  = list(contrast = "cubic",  orth = "gs"),
  "logphi-GS" = list(contrast = "logphi", orth = "gs"),
  "skew-GS"   = list(contrast = "skew",   orth = "gs")
)

methods_L <- lapply(method_spec,
                    function(m) run_fastica(U_w, m$contrast, m$orth,
                                            r = r, n_iter = 50))

# Objective belonging to each method, for ordering and labelling its panels.
obj_of <- function(m) contrast_defs[[method_spec[[m]]$contrast]]$obj

sapply(methods_L, dim)
```


## Comparing methods: matched factors {.tabset}

All six methods are aligned to a common reference, the polar cubic solution. For
each method the factors are matched to the reference by maximizing the total
absolute correlation (Hungarian algorithm), then sign-flipped so each matched
factor correlates positively with its reference factor. (Sign flipping is
probably unnecessary here because we are using asymmetric contrasts, but necessary in general.)

Factors are then ordered by how well the methods agree on them: the `mean`
column below is the average absolute correlation with the reference over the
five non-reference methods, and rows are sorted by it, so the factors all
methods recover go at the top and the contested ones at the bottom. Values near
1 mean a method recovers essentially the same factor as the reference; small
values mean the factor has no good counterpart. The reference column is 1 by
construction and is excluded from the mean.

```{r method-compare-align}
ref_name <- "cubic"
Lc       <- methods_L[[ref_name]]
Lref     <- Lc[, order(obj_of(ref_name)(Lc), decreasing = TRUE), drop = FALSE]

# Match columns of L to those of Lref so as to maximize total |correlation|,
# then flip signs so each matched factor correlates positively with reference.
match_to_ref <- function(L, Lref) {
  C    <- cor(Lref, L)
  pr   <- RcppHungarian::HungarianSolver(-abs(C))$pairs
  perm <- pr[order(pr[, 1]), 2]
  cc   <- C[cbind(seq_along(perm), perm)]
  s    <- sign(cc)
  s[s == 0] <- 1
  list(L = sweep(L[, perm, drop = FALSE], 2, s, "*"), cor = abs(cc))
}

aligned   <- lapply(methods_L, match_to_ref, Lref = Lref)
match_cor <- sapply(aligned, `[[`, "cor")

# Reorder factors so the best-agreeing ones come first. Agreement is the mean
# |cor| with the reference over the five non-reference methods (the reference
# column is identically 1 and so carries no information).
agree <- rowMeans(match_cor[, setdiff(colnames(match_cor), ref_name),
                            drop = FALSE])
ord   <- order(agree, decreasing = TRUE)

Lref      <- Lref[, ord, drop = FALSE]
aligned   <- lapply(aligned, function(a)
                    list(L = a$L[, ord, drop = FALSE], cor = a$cor[ord]))
match_cor <- match_cor[ord, , drop = FALSE]
agree     <- agree[ord]
rownames(match_cor) <- sprintf("f%02d", seq_len(nrow(match_cor)))
round(cbind(match_cor, mean = agree), 3)
```

```{r method-compare-fn}
# Rows = factors, best-agreeing first; columns = methods. The number in each
# panel is |cor| with the reference factor; the row strip carries the mean over
# non-reference methods.
plot_method_grid <- function(aligned, match_cor, agree, si = sample_info,
                             color_by = c("celltype", "donor")) {
  color_by <- match.arg(color_by)
  cvar  <- switch(color_by,
                  celltype = si$celltype,
                  donor    = sub("-.*", "", si$id))
  cname <- switch(color_by,
                  celltype = "Cell type",
                  donor    = "Donor")
  cell_order <- order(si$celltype)
  nc <- nrow(aligned[[1]]$L)
  nf <- ncol(aligned[[1]]$L)
  mn <- names(aligned)
  flev <- sprintf("f%02d (%.2f)", seq_len(nf), agree)

  df <- do.call(rbind, lapply(mn, function(m) {
    data.frame(
      rank    = rep(seq_len(nc), nf),
      grp     = rep(cvar[cell_order], nf),
      loading = as.vector(aligned[[m]]$L[cell_order, ]),
      factor  = factor(rep(flev, each = nc), levels = flev),
      method  = factor(m, levels = mn)
    )
  }))

  lab <- data.frame(
    factor = factor(rep(flev, times = length(mn)), levels = flev),
    method = factor(rep(mn, each = nf), levels = mn),
    lab    = sprintf("%.2f", as.vector(match_cor))
  )

  p <- ggplot(df, aes(x = rank, y = loading, color = grp)) +
    geom_point(size = 0.45, alpha = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.2) +
    geom_text(data = lab, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.15, vjust = 1.4, size = 1.9, color = "grey20",
              inherit.aes = FALSE) +
    facet_grid(factor ~ method, scales = "free_y") +
    labs(x = NULL, y = "Loading", color = cname) +
    theme_bw(base_size = 9) +
    theme(axis.text.x      = element_blank(),
          axis.ticks.x     = element_blank(),
          axis.text.y      = element_text(size = 5),
          strip.text.x     = element_text(size = 8),
          strip.text.y     = element_text(size = 6),
          strip.background = element_rect(fill = "grey90", color = NA),
          panel.spacing    = grid::unit(0.15, "lines"),
          legend.position  = "bottom",
          legend.title     = element_text(size = 11),
          legend.text      = element_text(size = 10),
          legend.key.size  = grid::unit(1.1, "lines")) +
    guides(color = guide_legend(override.aes = list(size = 4, alpha = 1),
                                nrow = 2))
  if (color_by == "celltype")
    p <- p + scale_color_manual(values = celltype_palette)
  p
}
```

### Cell type

```{r method-compare-ct, echo=FALSE, fig.width=13, fig.height=41, dpi=100, fig.retina=1}
plot_method_grid(aligned, match_cor, agree, color_by = "celltype")
```

### Donor

```{r method-compare-d, echo=FALSE, fig.width=13, fig.height=41, dpi=100, fig.retina=1}
plot_method_grid(aligned, match_cor, agree, color_by = "donor")
```
