---
title: "pancreas_celseq2_snmf_k40_ms"
author: "Matthew Stephens"
date: "2025-07-03"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(fastTopics)
library(Matrix)
library(ggplot2)
library(cowplot)
library(ggrepel)
```

## Introduction

I want to look at the semi-NMF results for the Pancreas celseq data (`fit_pancreas_celseq2_snmf_k40.R'), and particularly see how consistent the results are from two different subsets of the data.

First I read in the data and the fit results.

```{r}
load("../data/pancreas.RData")
set.seed(1)

# Select the CEL-seq2 data (Muraro et al, 2016).
# This should select 2,285 cells.
i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i,]
counts      <- counts[i,]

# Remove genes that are expressed in fewer than 10 cells.
x      <- colSums(counts > 0)
j      <- which(x > 9)
counts <- counts[,j]

# Compute the shifted log counts.
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
Y <- MatrixExtra::mapSparse(counts/(a*s),log1p)

#randomly divide rows of Y into 2
subset = sample(1:nrow(Y), nrow(Y)/2)

load("../output/pancreas_celseq2_snmf_k40.RData")

colnames(fit.snmf.k40.1$F_pm) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.2$F_pm) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))

colnames(fit.snmf.k40.1$F_psd) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.2$F_psd) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))

colnames(fit.snmf.k40.1$F_lfsr) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.2$F_lfsr) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))





colnames(fit.snmf.k40.1$L_pm) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.2$L_pm) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.1$L_psd) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.2$L_psd) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.1$L_lfsr) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))
colnames(fit.snmf.k40.2$L_lfsr) <- c("Baseline",paste0("GEP",1:(ncol(fit.snmf.k40.1$F_pm)-1)))


```

## Look at cross correlations

We compute the cross correlations of the factors from the two fits. 
A histogram shows that xcors exceeding about 0.5 are 
"above the noise". 
```{r}
xcor = cor(fit.snmf.k40.1$F_pm,fit.snmf.k40.2$F_pm)
hist(xcor)
```

And generally we see a 1-1 match at this threshold:
```{r}
image(xcor>0.5)
```



## Find matches between the two fits

For each row of xcor (GEPs in first fit) find the names of the columns (GEPs in second fit) whose correlation exceed 0.5.
The matches are those GEPs that have exactly one such match (a few have 2 other GEPs that are correlated >0.5 suggesting they are somewhat split).
```{r}
above_noise = apply(xcor,1,function(x) names(x)[x > 0.5])
matches = above_noise[lapply(above_noise,length)==1]
matches
```
  
  
## Structure Plots with matching GEPs
  
Now I do structure plots with just the matching GEPs (with the same colors to help see the matching).

```{r}
sample_info <- transform(sample_info,celltype = factor(celltype))
celltype <- sample_info$celltype
celltype <-
 factor(celltype,
        c("acinar","ductal","activated_stellate","quiescent_stellate",
          "endothelial","macrophage","mast","schwann","alpha","beta",
          "delta","gamma","epsilon"))


L.1 = fit.snmf.k40.1$L_pm
L.2 = fit.snmf.k40.2$L_pm

p1.match <- structure_plot(L.1[,names(matches)],grouping = celltype[subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")

p2.match <- structure_plot(L.2[,unlist(matches)],grouping = celltype[-subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")

plot_grid(p1.match,p2.match,ncol=1)
```

  
## Volcano plots of matching GEPs  

Here I define a function for "volcano plots", based on code from the `gbcd` package. I changed the y axis to show the "baseline" expression of each gene; I included negative y axis just to allow space for gene names which otherwise got very cramped.

```{r}
volcano_plot=function(res,GEP){
  pdat <- data.frame(gene = rownames(res$F_pm), 
                   lfc = res$F_pm[,GEP], 
                   base = res$F_pm[,"Baseline"], 
                   lfsr = res$F_lfsr[, GEP],
                   stringsAsFactors = FALSE)
  pdat <- transform(pdat, lfsr = cut(lfsr, c(-1, 0.001, 0.01, 0.05, Inf)))
  rows  <- with(pdat, which(!(abs(lfc) > quantile(abs(lfc), 0.998) )))
  pdat[rows, "gene"] <- ""
  p = ggplot(pdat, aes(x = lfc, y = base, color = lfsr, label = gene)) + ylim(-2,max(pdat$base)) + geom_point() + 
    geom_text_repel(color = "black", size = 2.5, segment.color = "black",
                  segment.size = 0.25, min.segment.length = 0,
                  max.overlaps = Inf, na.rm = TRUE) +
  scale_color_manual(values = c("coral", "orange", "gold", "deepskyblue")) +
  labs(x = "log-fold change", y = "Baseline") + 
  guides(colour = guide_legend(override.aes = list(size = 2))) + 
  theme(plot.title      = element_text(hjust = 0.5,size = 12),
        axis.text       = element_text(size = 10),
        axis.title      = element_text(size = 10),
        legend.title    = element_text(size = 12),
        legend.text     = element_text(size = 10),
        legend.position = "bottom") +
  ggtitle(paste0(GEP))
  return(p)
}
```

For each matching pair of GEPs, plot the volcano plots
next to one another for comparison. I exclude the baseline here.
```{r}
for(GEP.1 in names(matches)[-1]){
  print(GEP.1)
  GEP.2 <- matches[[GEP.1]]
  p.1 <- volcano_plot(fit.snmf.k40.1, GEP.1)
  p.2 <- volcano_plot(fit.snmf.k40.2, GEP.2)
  p = plot_grid(p.1, p.2,nrow=1)
  plot(p)
}
```

## Closer look at GEP12,GEP14

Despite being "matches" these two look pretty different so I thought I would look a bit closer. It turns out this one is just above the correlation 0.5 threshold.

```{r}
xcor["GEP14","GEP12"]
plot(fit.snmf.k40.1$F_pm[,"GEP14"],fit.snmf.k40.2$F_pm[,"GEP12"],
     xlab = "GEP14",ylab = "GEP12",
     main = paste0("xcor = ",round(xcor["GEP14","GEP12"],3)))
```

Here I plot the structure plot for just these GEPs (I have to show baseline as the code requires at least two topics I think?)
The distribution of the GEPs across cell types are similar, which is somewhat reassuring that it might be a meaningful signal of some kind, even if the driving genes are not very clear since they do not stand out greatly from the rest.
```{r}
p1.sub <- structure_plot(L.1[,c("Baseline","GEP14")],grouping = celltype[subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")

p2.sub <- structure_plot(L.2[,c("Baseline","GEP12")],grouping = celltype[-subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")

plot_grid(p1.sub,p2.sub,ncol=1)
```



