---
title: "pancreas_celseq2_ica_02"
author: "Matthew Stephens"
date: "2026-09-16"
output:
  workflowr::wflow_html:
    toc: true
    toc_float: true
    toc_depth: 3
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

## Introduction

A second look at ICA on the pancreas CEL-seq2 data, incorporating lessons from the first analysis (`pancreas_celseq2_ica.Rmd`). Key differences from the first analysis:

- Data are **centred** (column means subtracted) before whitening, as is standard for ICA.
- We run **rank-$r$ (symmetric) fastICA with $r = 25$** rather than many rank-1 starts, which is faster and avoids the post-hoc deduplication step.
- We compare three contrast functions chosen to favour one-sided/sparse sources: $G(x) = x^3$ (skewness), $G(x) = \log\Phi(\alpha x)$ (log-normal CDF), and $G(x) = x|x|$ (the tilting term of TLC without the log-cosh).

```{r load-data}
library(fastICA)
library(Matrix)
library(ggplot2)
library(caret)

load("../data/pancreas.RData")
set.seed(1)

i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i, ]
counts      <- counts[i, ]

x      <- colSums(counts > 0)
j      <- which(x > 9)
counts <- counts[, j]

a <- 1
s <- rowSums(counts)
s <- s / mean(s)
Y <- MatrixExtra::mapSparse(counts / (a * s), log1p)
Y <- scale(Y, scale = FALSE)   # centre columns
```

## Preprocessing: SVD of centred data

```{r svd}
n      <- nrow(Y)
n.comp <- 25
r      <- n.comp

Y.svd <- svd(Y, nu = n.comp, nv = 0)
# U: n x n.comp, orthonormal columns (left singular vectors of centred Y)
U_c <- Y.svd$u[, 1:n.comp]
# Whitened row space (n.comp x n, as used by rank-r fastICA)
U_w <- sqrt(n) * t(U_c)
```

## Helper functions

```{r helpers}
celltype_palette <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
  "#A65628", "#F781BF", "#1B9E77", "#D95F02", "#7570B3",
  "#E7298A", "#66A61E", "#E6AB02", "#A6761D", "#666666"
)

# Polar factor via eigendecomposition (avoids SVD column reordering).
polar <- function(W) {
  eig      <- eigen(t(W) %*% W, symmetric = TRUE)
  Ainvhalf <- eig$vectors %*%
               diag(1 / sqrt(pmax(eig$values, 1e-14))) %*%
               t(eig$vectors)
  W %*% Ainvhalf
}

prune_and_count_cluster <- function(L, tau = 0.9) {
  L        <- L[, apply(L, 2, sd) > 1e-10, drop = FALSE]
  dist_mat <- as.dist(1 - abs(cor(L)))
  hc       <- hclust(dist_mat, method = "complete")
  clusters <- cutree(hc, h = 1 - tau)
  kept_indices        <- match(unique(clusters), clusters)
  cluster_sizes_table <- table(clusters)
  kept_cluster_ids    <- clusters[kept_indices]
  cluster_sizes       <- as.integer(cluster_sizes_table[as.character(kept_cluster_ids)])
  list(pruned_matrix      = L[, kept_indices, drop = FALSE],
       cluster_sizes      = cluster_sizes,
       kept_indices       = kept_indices,
       cluster_assignments = clusters)
}

plot_loadings_matrix <- function(Lhat, label_prefix = "c:", si = sample_info,
                                  obj_fn = function(L) colMeans(L^4),
                                  max_panels = 25, labels = NULL) {
  obj    <- obj_fn(Lhat)
  n_comp <- ncol(Lhat)
  if (is.null(labels)) {
    o           <- order(obj)
    Lhat        <- Lhat[, o, drop = FALSE]
    obj         <- obj[o]
    comp_labels <- make.unique(paste0(label_prefix, seq_len(n_comp),
                                      " obj:", round(obj, 3)))
  } else {
    comp_labels <- make.unique(paste0(labels, " obj:", round(obj, 3)))
  }
  cell_order <- order(si$celltype)
  df <- data.frame(
    rank      = rep(seq_len(nrow(Lhat)), n_comp),
    celltype  = rep(si$celltype[cell_order], n_comp),
    loading   = as.vector(Lhat[cell_order, ]),
    component = factor(rep(comp_labels, each = nrow(Lhat)),
                       levels = comp_labels)
  )
  pages <- split(comp_labels,
                 ceiling(seq_along(comp_labels) / max_panels))
  for (pg in seq_along(pages)) {
    pg_labels <- pages[[pg]]
    df_pg     <- df[df$component %in% pg_labels, ]
    df_pg$component <- factor(as.character(df_pg$component),
                               levels = pg_labels)
    pg_title <- if (length(pages) > 1)
      paste0("page ", pg, "/", length(pages)) else ""
    p <- ggplot(df_pg, aes(x = rank, y = loading, color = celltype)) +
      geom_point(size = 0.5, alpha = 0.7) +
      geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
      facet_wrap(~ component, ncol = 5, scales = "free_y") +
      scale_color_manual(values = celltype_palette) +
      labs(x = NULL, y = "Loading", color = "Cell type", title = pg_title) +
      theme_bw(base_size = 10) +
      theme(axis.text.x  = element_blank(),
            axis.ticks.x = element_blank(),
            strip.text   = element_text(size = 7,
                                        margin = margin(2, 0, 2, 0)),
            strip.background = element_rect(fill = "grey90", color = NA),
            legend.position  = "bottom")
    print(p)
  }
}
```

## Contrast functions

```{r contrasts}
# --- G(x) = x^3 (skewness) ---
# g(x) = G'(x) = 3x^2,  g'(x) = G''(x) = 6x
fastica_update_cubic_rankr <- function(U, W) {
  P  <- t(U) %*% W
  G  <- 3 * P^2
  G2 <- 6 * P
  W  <- U %*% G - sweep(W, 2, colSums(G2), "*")
  polar(W)
}
objective_cubic <- function(L) colMeans(L^3)

# --- log-Phi ---
# G(x) = log Phi(alpha*x),  g = alpha*h,  g' = alpha^2*(-alpha*x*h - h^2)
fastica_update_logphi_rankr <- function(U, W, alpha = 2) {
  P  <- t(U) %*% W
  u  <- alpha * P
  h  <- exp(dnorm(u, log = TRUE) - pnorm(u, log.p = TRUE))
  G  <- alpha * h
  G2 <- alpha^2 * (-u * h - h^2)
  W  <- U %*% G - sweep(W, 2, colSums(G2), "*")
  polar(W)
}
objective_logphi <- function(L, alpha = 2) colMeans(pnorm(alpha * L, log.p = TRUE))

# --- x|x| (skew tilt, no log-cosh) ---
# g(x) = 2|x|,  g'(x) = 2*sign(x),  G(x) = x|x|
fastica_update_skew_rankr <- function(U, W) {
  P  <- t(U) %*% W
  G  <- 2 * abs(P)
  G2 <- 2 * sign(P)
  W  <- U %*% G - sweep(W, 2, colSums(G2), "*")
  polar(W)
}
objective_skew <- function(L) colMeans(L * abs(L))

# Gram-Schmidt reorthogonalization ordered by decreasing objective.
# W: n.comp x r matrix with orthonormal columns.
# L: n x r loadings matrix (= t(U_w) %*% W).
# Returns W with orthonormal columns, best-objective direction first.
gs_reorth_W <- function(W, L, obj_fn) {
  ord <- order(obj_fn(L), decreasing = TRUE)
  V   <- W[, ord, drop = FALSE]
  for (j in seq_len(ncol(V))) {
    if (j > 1)
      for (l in seq_len(j - 1))
        V[, j] <- V[, j] - sum(V[, j] * V[, l]) * V[, l]
    nrm <- sqrt(sum(V[, j]^2))
    if (nrm > 1e-10) V[, j] <- V[, j] / nrm
  }
  V
}
```

## Rank-r fastICA: cubic contrast (x^3)

```{r ica-cubic}
set.seed(1)
n_iter <- 500
W_cubic <- polar(matrix(rnorm(n.comp * r), n.comp, r))
for (i in seq_len(n_iter))
  W_cubic <- fastica_update_cubic_rankr(U_w, W_cubic)

Lhat_cubic <- t(U_w) %*% W_cubic   # n x r
plot_loadings_matrix(Lhat_cubic, label_prefix = "cub:",
                     obj_fn = objective_cubic)
```

## Rank-r fastICA: log-Phi contrast (alpha = 2)

```{r ica-logphi}
set.seed(1)
n_iter <- 500
W_lp <- polar(matrix(rnorm(n.comp * r), n.comp, r))
for (i in seq_len(n_iter))
  W_lp <- fastica_update_logphi_rankr(U_w, W_lp, alpha = 2)

Lhat_lp <- t(U_w) %*% W_lp
plot_loadings_matrix(Lhat_lp, label_prefix = "lp:",
                     obj_fn = objective_logphi)
```

## Rank-r fastICA: x|x| contrast

```{r ica-skew}
set.seed(1)
n_iter <- 500
W_skew <- polar(matrix(rnorm(n.comp * r), n.comp, r))
for (i in seq_len(n_iter))
  W_skew <- fastica_update_skew_rankr(U_w, W_skew)

Lhat_skew <- t(U_w) %*% W_skew
plot_loadings_matrix(Lhat_skew, label_prefix = "sk:",
                     obj_fn = objective_skew)
```

## Gram-Schmidt versions

Each iteration applies the contrast update followed by Gram-Schmidt reorthogonalization (instead of polar). Components are ordered by decreasing objective before GS, so the best-objective direction is processed first and kept exactly; subsequent components are deflated against all preceding ones. After GS the columns remain orthonormal.

### GS: cubic contrast (x^3)

```{r ica-cubic-gs}
set.seed(1)
n_iter <- 500
W_cubic_gs <- polar(matrix(rnorm(n.comp * r), n.comp, r))
for (i in seq_len(n_iter)) {
  P              <- t(U_w) %*% W_cubic_gs
  G              <- 3 * P^2
  G2             <- 6 * P
  W_cubic_gs     <- U_w %*% G - sweep(W_cubic_gs, 2, colSums(G2), "*")
  L_tmp          <- t(U_w) %*% W_cubic_gs
  W_cubic_gs     <- gs_reorth_W(W_cubic_gs, L_tmp, objective_cubic)
}

Lhat_cubic_gs <- t(U_w) %*% W_cubic_gs
plot_loadings_matrix(Lhat_cubic_gs, label_prefix = "cub-gs:",
                     obj_fn = objective_cubic)
```

### GS: log-Phi contrast (alpha = 2)

```{r ica-logphi-gs}
set.seed(1)
n_iter <- 500
alpha  <- 2
W_lp_gs <- polar(matrix(rnorm(n.comp * r), n.comp, r))
for (i in seq_len(n_iter)) {
  P       <- t(U_w) %*% W_lp_gs
  u       <- alpha * P
  h       <- exp(dnorm(u, log = TRUE) - pnorm(u, log.p = TRUE))
  G       <- alpha * h
  G2      <- alpha^2 * (-u * h - h^2)
  W_lp_gs <- U_w %*% G - sweep(W_lp_gs, 2, colSums(G2), "*")
  L_tmp   <- t(U_w) %*% W_lp_gs
  W_lp_gs <- gs_reorth_W(W_lp_gs, L_tmp, objective_logphi)
}

Lhat_lp_gs <- t(U_w) %*% W_lp_gs
plot_loadings_matrix(Lhat_lp_gs, label_prefix = "lp-gs:",
                     obj_fn = objective_logphi)
```

### GS: x|x| contrast

```{r ica-skew-gs}
set.seed(1)
n_iter   <- 500
W_skew_gs <- polar(matrix(rnorm(n.comp * r), n.comp, r))
for (i in seq_len(n_iter)) {
  P         <- t(U_w) %*% W_skew_gs
  G         <- 2 * abs(P)
  G2        <- 2 * sign(P)
  W_skew_gs <- U_w %*% G - sweep(W_skew_gs, 2, colSums(G2), "*")
  L_tmp     <- t(U_w) %*% W_skew_gs
  W_skew_gs <- gs_reorth_W(W_skew_gs, L_tmp, objective_skew)
}

Lhat_skew_gs <- t(U_w) %*% W_skew_gs
plot_loadings_matrix(Lhat_skew_gs, label_prefix = "sk-gs:",
                     obj_fn = objective_skew)
```
