---
title: "Automatic Grouping of Factors: Pancreas"
author: "Ziang Zhang"
date: "2025-02-12"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---


Let's try some automatic grouping of factors for the pancreas dataset, based on the grouping information provided in the metadata.

```{r}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)
set.seed(1)
subsample_cell_types <- function (x, n = 1000) {
  cells <- NULL
  groups <- levels(x)
  for (g in groups) {
    i  <-  which(x == g)
    n0 <- min(n,length(i))
    i  <- sample(i,n0)
    cells <- c(cells,i)
  }
  return(sort(cells))
}
load("../data/pancreas.RData")
load("../output/pancreas_factors.RData")
load("../output/pancreas_factors2.RData")
```

```{r}
source("../code/group_factors.R")
```

### A Simple Automatic Approach Based on ANOVA

Let \(l_{ki}\) denote the loading of observation \(i\) on the $k$th factor, and let \(G_i\) denote the grouping information (e.g. cell-type) for observation (e.g. cell) \(i\). 
Suppose \(G_i\) can take values in \(\{g_1, g_2, \ldots, g_L\}\) (e.g. $L$ possible cell-types). 

A straightforward way to identify group-specific factors is to perform an ANOVA on each factor's loadings. 
Specifically, for the $k$th factor, we regress the loadings \(\boldsymbol{l}_k\) on the set of indicator variables \(\{\mathbb{I}(G_i = c_l)\}_{l=1}^L\):

\[
l_{ki} \sim \sum_{l=1}^L \beta_l \,\mathbb{I}(G_i = c_l).
\]

From the regression model for each factor, we obtain a statistic \(S_k\) that measures how strongly the grouping structure explains variation in the \(k\)-th factor's loadings. 
For instance, \(S_k\) could be the \(R^2\) of this regression model or the \(F\)-statistic from the ANOVA test. 
After computing \(\{S_k\}_{k=1}^K\), we can rank the factors by their relevance to the grouping, then choose a threshold (for example, by examining an elbow plot) to automatically select the ``group-specific'' factors.

Below, let's apply this approach to the pancreas dataset, using \(R^2\) as the relevance statistic.



### flashier NMF

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
cells <- subsample_cell_types(sample_info$celltype,n = 500)
L <- fl_nmf_ldf$L
k <- ncol(L)
colnames(L) <- paste0("k",1:k)
```

Take a look at the elbow plot:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
ordered_df_tech <- ANOVA_factors(L[cells,], sample_info$tech[cells], stats = "R2")
ordered_df_celltype <- ANOVA_factors(L[cells,], sample_info$celltype[cells], stats = "R2")
par(mfrow = c(2,1))
plot(ordered_df_tech$rank, ordered_df_tech$stats, type = "o", xlab = "Rank", ylab = "R2", main = "batchtypes", ylim = c(0,1.1))
text(ordered_df_tech$rank, ordered_df_tech$stats, labels = ordered_df_tech$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
plot(ordered_df_celltype$rank, ordered_df_celltype$stats, type = "o", xlab = "Rank", ylab = "R2", main = "celltypes", ylim = c(0,1.1))
text(ordered_df_celltype$rank, ordered_df_celltype$stats, labels = ordered_df_celltype$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
par(mfrow = c(1,1))
```



Cut off the factors with $R^2$ less than 0.7:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.7, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.7, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```


Cut off the factors with $R^2$ less than 0.5:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.5, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.5, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```


Try reduce to the cut-off to 0.3:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.3, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.3, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```



### NMF (NNLM)

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
W <- nmf$W
k <- ncol(W)
d <- apply(W,2,max)
scale_cols <- function (A, b)
  t(t(A) * b)
W <- scale_cols(W,1/d)
colnames(W) <- paste0("k",1:k)
```

Take a look at the elbow plot:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
ordered_df_tech <- ANOVA_factors(W[cells,], sample_info$tech[cells], stats = "R2")
ordered_df_celltype <- ANOVA_factors(W[cells,], sample_info$celltype[cells], stats = "R2")
par(mfrow = c(2,1))
plot(ordered_df_tech$rank, ordered_df_tech$stats, type = "o", xlab = "Rank", ylab = "R2", main = "batchtypes", ylim = c(0,1.1))
text(ordered_df_tech$rank, ordered_df_tech$stats, labels = ordered_df_tech$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
plot(ordered_df_celltype$rank, ordered_df_celltype$stats, type = "o", xlab = "Rank", ylab = "R2", main = "celltypes", ylim = c(0,1.1))
text(ordered_df_celltype$rank, ordered_df_celltype$stats, labels = ordered_df_celltype$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
par(mfrow = c(1,1))
```


Cut off the factors with $R^2$ less than 0.7:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = W[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.7, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = W[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.7, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```


Cut off the factors with $R^2$ less than 0.5:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = W[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.5, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = W[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.5, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```

Cut off the factors with $R^2$ less than 0.3:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = W[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.3, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = W[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.3, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```



### Topic model (fastTopics)

```{r}
L <- poisson2multinom(pnmf)$L
```

Take a look at the elbow plot:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
ordered_df_tech <- ANOVA_factors(L[cells,], sample_info$tech[cells], stats = "R2")
ordered_df_celltype <- ANOVA_factors(L[cells,], sample_info$celltype[cells], stats = "R2")
par(mfrow = c(2,1))
plot(ordered_df_tech$rank, ordered_df_tech$stats, type = "o", xlab = "Rank", ylab = "R2", main = "batchtypes", ylim = c(0,1.1))
text(ordered_df_tech$rank, ordered_df_tech$stats, labels = ordered_df_tech$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
plot(ordered_df_celltype$rank, ordered_df_celltype$stats, type = "o", xlab = "Rank", ylab = "R2", main = "celltypes", ylim = c(0,1.1))
text(ordered_df_celltype$rank, ordered_df_celltype$stats, labels = ordered_df_celltype$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
par(mfrow = c(1,1))
```


Cut off the factors with $R^2$ less than 0.7:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.7, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.7, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```

Cut off the factors with $R^2$ less than 0.5:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.5, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.5, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```


Cut off the factors with $R^2$ less than 0.3:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.3, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.3, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```



### Flashier semi-NMF

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
L <- fl_snmf_ldf$L
x <- apply(L,2,function (x) quantile(x,0.995))
L <- scale_cols(L,1/x)
colnames(L) <- paste0("k",1:k)
```

Take a look at the elbow plot:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=8, fig.width=8}
ordered_df_tech <- ANOVA_factors(W[cells,], sample_info$tech[cells], stats = "R2")
ordered_df_celltype <- ANOVA_factors(W[cells,], sample_info$celltype[cells], stats = "R2")
par(mfrow = c(2,1))
plot(ordered_df_tech$rank, ordered_df_tech$stats, type = "o", xlab = "Rank", ylab = "R2", main = "batchtypes", ylim = c(0,1.1))
text(ordered_df_tech$rank, ordered_df_tech$stats, labels = ordered_df_tech$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
plot(ordered_df_celltype$rank, ordered_df_celltype$stats, type = "o", xlab = "Rank", ylab = "R2", main = "celltypes", ylim = c(0,1.1))
text(ordered_df_celltype$rank, ordered_df_celltype$stats, labels = ordered_df_celltype$factor, pos = 3, cex = 0.6)
abline(h = c(0.3,0.5,0.7), col = "red", lty = 2)
par(mfrow = c(1,1))
```


Cut off the factors with $R^2$ less than 0.7:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.7, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.7, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```


Cut off the factors with $R^2$ less than 0.5:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.5, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.5, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```

Cut off the factors with $R^2$ less than 0.3:

```{r, results="hide", message=FALSE, warning=FALSE, fig.height=6, fig.width=8}
p1 <- structure_plot_group(L = L[cells,], group_vec = sample_info$tech[cells],
                           cutoff = 0.3, stats = "R2", group_name = "data-set")
p2 <- structure_plot_group(L = L[cells,], group_vec = sample_info$celltype[cells],
                           cutoff = 0.3, stats = "R2", group_name = "cell-type")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```


