---
title: Preparation and initial exploration of the LSA 2021 pancreas data set
author: Eric Weine and Peter Carbonetto
output: workflowr::wflow_html
---

Here we will prepare the single-cell RNA-seq data from
[Stancill et al 2021][stancill_2021_lsa] for analysis with the topic
model and other matrix factorization methods. The data files were
obtained by downloading and the `GSE156175_RAW.tar` file from
[GEO accession GSE156175][stancill_2021_lsa_geo]. To run this script,
download this file from GEO and extract the contents into a folder
called "GSE156175_RAW" in the data directory.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the R packages used to perform the data processing and initial
analyses:

```{r load-pkgs, message=FALSE}
library(Matrix)
library(readr)
library(fastTopics)
library(dplyr)
library(Seurat)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Import the sample information:

```{r import-barcodes, message=FALSE}
barcodes1<-read_tsv("../data/GSE156175_RAW/GSM5842388_Rep1_S1_barcodes.tsv.gz",
                    col_names = "barcode")
barcodes2 <- read_tsv("../data/GSE156175_RAW/GSM4726017_S2_barcodes.tsv.gz",
                      col_names = "barcode")
barcodes3 <- read_tsv("../data/GSE156175_RAW/GSM4726018_S3_barcodes.tsv.gz",
                      col_names = "barcode")
barcodes4 <- read_tsv("../data/GSE156175_RAW/GSM4726019_S4_barcodes.tsv.gz",
                      col_names = "barcode")
barcodes1$barcode <- substr(barcodes1$barcode,1,nchar(barcodes1$barcode) - 2)
barcodes2$barcode <- substr(barcodes2$barcode,1,nchar(barcodes2$barcode) - 2)
barcodes3$barcode <- substr(barcodes3$barcode,1,nchar(barcodes3$barcode) - 2)
barcodes4$barcode <- substr(barcodes4$barcode,1,nchar(barcodes4$barcode) - 2)
barcodes1$condition <- "Untreated"
barcodes2$condition <- "IL-1B"
barcodes3$condition <- "IFNg"
barcodes4$condition <- "IL-1B_IFNg"
barcodes <- rbind(barcodes1, barcodes2, barcodes3, barcodes4)
barcodes$cell_bc <- paste0(barcodes$barcode, "_", barcodes$condition)
barcodes$condition <- factor(barcodes$condition)
```

Import the gene information:

```{r import-gene-info, message=FALSE}
genes1 <- read_tsv("../data/GSE156175_RAW/GSM5842388_Rep1_S1_features.tsv.gz",
                   col_names = c("ensembl", "symbol", "type"))
genes2 <- read_tsv("../data/GSE156175_RAW/GSM4726017_S2_features.tsv.gz",
                   col_names = c("ensembl", "symbol", "type"))
genes3 <- read_tsv("../data/GSE156175_RAW/GSM4726018_S3_features.tsv.gz",
                   col_names = c("ensembl", "symbol", "type"))
genes4 <- read_tsv("../data/GSE156175_RAW/GSM4726019_S4_features.tsv.gz",
                   col_names = c("ensembl", "symbol", "type"))
genes <- genes1
genes$type <- factor(genes$type)
```

Now import in the read counts:

```{r import-counts, message=FALSE}
counts1 <- t(readMM("../data/GSE156175_RAW/GSM5842388_Rep1_S1_matrix.mtx.gz"))
counts2 <- t(readMM("../data/GSE156175_RAW/GSM4726017_S2_matrix.mtx.gz"))
counts3 <- t(readMM("../data/GSE156175_RAW/GSM4726018_S3_matrix.mtx.gz"))
counts4 <- t(readMM("../data/GSE156175_RAW/GSM4726019_S4_matrix.mtx.gz"))
rownames(counts1) <- paste0(barcodes1$barcode, "_", barcodes1$condition)
rownames(counts2) <- paste0(barcodes2$barcode, "_", barcodes2$condition)
rownames(counts3) <- paste0(barcodes3$barcode, "_", barcodes3$condition)
rownames(counts4) <- paste0(barcodes4$barcode, "_", barcodes4$condition)
colnames(counts1) <- genes1$symbol
colnames(counts2) <- genes2$symbol
colnames(counts3) <- genes3$symbol
colnames(counts4) <- genes4$symbol
counts <- rbind(
  as.matrix(counts1),
  as.matrix(counts2),
  as.matrix(counts3),
  as.matrix(counts4)
)
counts <- as(counts, "CsparseMatrix")
```

Removed genes that are not expressed in any cell:

```{r filter-genes-1}
x      <- colSums(counts)
j      <- which(x > 0)
genes  <- genes[j,]
counts <- counts[,j]
```

Remove cells in terms (i) total UMI count and (ii) very few
expressed genes:

```{r filter-cells}
x        <- rowSums(counts > 0)
i        <- which(x > 2000)
barcodes <- barcodes[i,]
counts   <- counts[i,]
x        <- rowSums(counts)
i        <- which(x > 60000)
barcodes <- barcodes[i,]
counts   <- counts[i,]
```

PC: The first filtering step here (x > 2,000) seems a bit
aggressive. I think we could have retained some more cells.

Remove cells with high mitochondrial counts, and remove *MALAT1*,
ribosomal genes, and mitochondrial genes for further analysis:

```{r filter-genes-2, eval=FALSE}
mito_genes <- which(substr(colnames(counts),1,2) == "mt")
s          <- rowSums(counts)
s_mito     <- counts[,mito_genes]
prop_mito  <- rowSums(s_mito)/s
i          <- which(prop_mito < 0.1)
counts  <- counts[i,]
j <- which(!(grepl("^mt-",colnames(counts)) |
               grepl("^Rp[sl]",colnames(counts))))
counts <- counts[,j]
j      <- which(colnames(counts) != "Malat1")
counts <- counts[,j]
j      <- which(colSums(counts > 0) > 2)
counts <- counts[,j]
barcodes <- barcodes %>% filter(cell_bc %in% rownames(counts))
```

Fit a topic model with $K = 12$.

```{r, eval=FALSE}
set.seed(1)
tm_fit <- fit_poisson_nmf(
  X = as(counts, "CsparseMatrix"),
  k = 12,
  control = list(nc = 7)
)
```

Here is the initial structure plot:

```{r, eval=FALSE}
structure_plot(tm_fit, n = Inf)
```

Below is a heatmap annotation of the marker genes given in the
Stancill paper:

```{r, eval=FALSE}
scale_rows <- function (A)
  A / apply(A,1,max)
# using marker genes, I would like to connect each topic to a celltype
marker_genes <- c("Ins1","Ins2","Gcg","Sst",
                  "Ppy","Krt17","Ccr5",
                  "Col1a1", "Cd34", "Cpa1")
F <- poisson2multinom(tm_fit)$F
F <- F[marker_genes,]
F <- scale_rows(F)
rownames(F) <- marker_genes
p <- annotation_heatmap(F,select_features = "all",verbose = FALSE)
print(p)
```

Below are some comments relating my guesses about factor
associations. Many of the beta associated factors are non-specific:

```{r, eval=FALSE}
# factor associations
# k1: ?
# k2: Acinar
# k3: Endothelial/Mesnchymal
# k4: ? (presumably Beta)
# k5: Ductal
# k6: Beta
# k7: ? (presumably Beta)
# k8: Delta
# k9: Alpha
# k10: Gamma (PP) 
# k11: Macrophage
# k12: ? (presumably Beta)
```

Now, I do celltype assignments. I used 1/3 as a threshold on the topic
proportions to determine celltype because this seemed to look
reasonable visually. Below is the structure plot with cells grouped by
type.

```{r, eval=FALSE}
L <- poisson2multinom(tm_fit)$L
L_df <- as.data.frame(L)
L_df <- L_df %>%
  dplyr::mutate(
    cluster = case_when(
      k2 > (1/3) ~ "Acinar",
      k3 > (1/3) ~ "Endothelial/Mesnchymal",
      k5 > (1/3) ~ "Ductal",
      k8 > (1/3) ~ "Delta",
      k9 > (1/3) ~ "Alpha",
      k10 > (1/3) ~ "Gamma",
      k11 > (1/3) ~ "Macrophage",
      TRUE ~ "Beta"
    )
  )
structure_plot(tm_fit, grouping = L_df$cluster, gap = 20, n = Inf)
```

Overall, this clustering looks quite good, but there is a small pocket
of beta cells that have loadings on the green, black, and grey
factors, which indicate that they could also be alpha or delta
cells. Perhaps additional investigation into these cells (e.g. with a
UMAP) could help distinguish their precise types. Or, we could exclude
them as not having a clear type.

Next, I investigated a UMAP plot to see how our clusters looked. Below
is an elbow plot of the PCs created by Seurat

```{r, eval=FALSE}
barcodes$celltype <- L_df$cluster
counts <- counts[,!duplicated(colnames(counts))]
so <- CreateSeuratObject(counts = Matrix::t(counts), meta.data = barcodes)
so <- SCTransform(so, verbose = FALSE)
so <- RunPCA(so, verbose = FALSE)
ElbowPlot(so)
```

```{r, eval=FALSE}
so <- RunUMAP(so, dims = 1:10, verbose = FALSE)
DimPlot(so, group.by = "celltype")
```

There are clearly a few cells on the UMAP plot that appear to lie in a
different cluster than their assignment. However, I'm not convinced
that this plot is any more reliable than the topic model structure
plot.

[stancill_2021_lsa]: https://www.life-science-alliance.org/content/4/6/e202000949
[stancill_2021_lsa_geo]: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE156175
