---
title: NMF analysis of the "pancreas cytokine" data set (untreated mouse only)
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform a NMF analyses of the
["pancreas cytokine" data set](pancreas_cytokine.html), focussing on
the scRNA-seq data from untreated mouse only.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load packages used to process the data, perform the analyses, and
create the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(flashier)
library(singlecelljamboreeR)
library(ggplot2)
library(cowplot)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the prepared data set:

```{r load-data}
load("../data/pancreas_cytokine.RData")
```

Here we will analyze the cells from the untreated mouse only:

```{r filter-cells-1}
i       <- which(samples$mouse == "S1")
samples <- samples[i,]
counts  <- counts[i,]
```

Remove two cells that appear to be outliers:

```{r filter-cells-2}
outliers <- c("TTTGTTGTCGTTAGTG-1","TTTGTTGGTAGAGCTG-1")
i        <- which(!is.element(samples$barcode,outliers))
samples  <- samples[i,]
counts   <- counts[i,]
```

Remove genes that are expressed in fewer than 5 cells:

```{r filter-genes}
j      <- which(colSums(counts > 0) > 4)
genes  <- genes[j,]
counts <- counts[,j]
```

This is the dimension of the data set we will analyze:

```{r dim-counts}
dim(counts)
```

For the Gaussian-based analyses, we will need the shifted log counts:

```{r shifted-log-counts}
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
shifted_log_counts <- log1p(counts/(a*s))
rownames(shifted_log_counts) <- NULL
```

## Topic model (fastTopics)

Fit a topic model with $K = 7$ topics to the counts:

```{r fit-topic-model, cache=TRUE}
tm <- fit_poisson_nmf(counts,k = 7,init.method = "random",method = "em",
                      numiter = 40,verbose = "none",
					  control = list(numiter = 4,nc = 8,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 40,
                      control = list(numiter = 4,nc = 8,extrapolate = TRUE),
					  verbose = "none")
```

Structure plot comparing the topics to the clusters:

```{r structure-plot-topic-model, fig.height=2.25, fig.width=6.5, results="hide", message=FALSE}
topic_colors <- c("gainsboro","darkorange","darkblue","forestgreen",
                  "dodgerblue","gold","red")
L <- poisson2multinom(tm)$L
clusters <- as.character(samples$cluster)
clusters[clusters == "islet"]                 <- "beta"
clusters[clusters == "beta" & L[,"k6"] > 0.3] <- "alpha"
clusters[clusters == "beta" & L[,"k7"] > 0.2] <- "delta+gamma"
clusters[clusters == "beta" & L[,"k4"] > 0.2] <- "beta(Ins1-)"
clusters <- factor(clusters)
structure_plot(L,grouping = clusters,gap = 10,colors = topic_colors,
               topics = 1:7)
```

Based on the estimated $\mathbf{F}$, we have the following potential
interpretation of the topics:

+ k5 = $\beta$ cells (*Ins1*, *Ins2*, *Mafa*)
+ k4 = *Ins1-* $\beta$ cells?
+ k6 = $\alpha$ cells (*Gcg*, *Mafb*)
+ k7 = $\delta$, $\gamma$ and $\epsilon$ cells (*Sst*, *Ghrl*)
+ k2 = a non-$\alpha$ islet process
+ k3 = macrophages, endothelial cells, mesenchymal cells and duct cells

```{r annotation-heatmap-topic-model, fig.height=3, fig.width=3}
scale_rows <- function (A)
  A / apply(A,1,max)
marker_genes <- c("Ins1","Ins2","Mafa","Gcg","Mafb","Sst","Ghrl",
                  "Ppy","Chga","Iapp","Krt19","Ccr5","Pecam1","Esam",
				  "Col1a1","Ghrl")
j <- match(marker_genes,genes$symbol)
F <- poisson2multinom(tm)$F
F <- F[j,]
F <- scale_rows(F)
rownames(F) <- marker_genes
topics <- paste0("k",c(5,4,6,7,2,3))
annotation_heatmap(F[,topics],select_features = "all",verbose = FALSE)
```

Let's now try with $K = 13$ topics:

```{r fit-topic-model-k-13, cache=TRUE}
tm <- fit_poisson_nmf(counts,k = 13,init.method = "random",method = "em",
                      numiter = 40,verbose = "detailed",
                      control = list(numiter = 4,nc = 8,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 40,
                      control = list(numiter = 4,nc = 8,extrapolate = TRUE),
                      verbose = "detailed")
```

ADD TEXT HERE:

```{r structure-plot-topic-model-k-13, fig.height=2.25, fig.width=6.5, results="hide", message=FALSE}
L <- poisson2multinom(tm)$L
clusters <- samples$cluster
# clusters <- as.character(samples$cluster)
# clusters[clusters == "islet"] <- "beta"
# clusters[clusters == "beta" & L[,"k13"] > 0.25] <- "alpha"
# clusters[clusters == "beta" & L[,"k3"] > 0.25] <- "delta"
# clusters[clusters == "beta" & L[,"k6"] > 0.25] <- "gamma"
# clusters <- factor(clusters,
#                    c("endothelial-mesenchymal","macrophage","duct",
#                      "beta","alpha","delta","gamma"))
# saveRDS(clusters,"clusters.rds")
structure_plot(tm,grouping = clusters,gap = 10)
```

## EBNMF (flashier)

Next fit an NMF to the shifted log counts using flashier, also with $K
= 7$:

```{r flashier-nmf, cache=TRUE, warning=FALSE}
n  <- nrow(samples)
x  <- rpois(1e7,1/n)
s1 <- sd(log(x + 1))
fl_nmf <- flash(shifted_log_counts,S = s1,ebnm_fn = ebnm_point_exponential,
                var_type = 2,greedy_Kmax = 7,backfit = FALSE,
				nullcheck = FALSE,verbose = 0)
fl_nmf <- flash_backfit(fl_nmf,extrapolate = FALSE,maxiter = 40,verbose = 0)
fl_nmf <- flash_backfit(fl_nmf,extrapolate = TRUE,maxiter = 80,verbose = 0)
```

Structure plot comparing the factors to the clusters:

```{r structure-plot-flashier-nmf, fig.height=2.5, fig.width=6.5, results="hide", message=FALSE}
L <- ldf(fl_nmf,type = "i")$L
colnames(L) <- paste0("k",1:7)
clusters <- as.character(samples$cluster)
clusters[clusters == "endothelial-mesenchymal"] <- "mesen.+endothelial"
clusters[clusters == "mesen.+endothelial" & L[,"k6"] > 0.5] <- "mesenchymal"
clusters[clusters == "islet"] <- "beta"
clusters[clusters == "beta" & L[,"k3"] > 0.5]  <- "alpha+delta+gamma"
clusters[clusters == "beta" & L[,"k7"] > 0.25] <- "beta(Ins1-)"
clusters <- factor(clusters)
structure_plot(L[,-1],grouping = clusters,gap = 10,colors = topic_colors[-1]) +
  labs(y = "membership")
```

Possible interpretation of the factors:

+ k4 = $\beta$ cells (*Ins1*, *Ins2*, *Mafa*)
+ k7 = *Ins1-* $\beta$ cells?
+ k3 = $\alpha$, $\delta$ and $\gamma$ cells
+ k5 = macrophages (*Ccr5*)
+ k2 = endothelial cells (*Pecam1*, *Esam*)
+ k6 = mesenchymal cells (*Col1a1*)

```{r annotation-heatmap-flashier-nmf, fig.height=3, fig.width=3}
scale_cols <- function (A) {
  b <- apply(A,2,max)
  return(t(t(A) * b))
}
marker_genes <- c("Ins1","Ins2","Mafa","Gcg","Mafb","Sst","Ghrl",
                  "Ppy","Chga","Iapp","Krt19",
                  "Ccr5","Pecam1","Esam","Col1a1","Ghrl")
j <- match(marker_genes,genes$symbol)
F <- ldf(fl_nmf,type = "i")$F
F <- scale_cols(F)
F <- F[j,]
rownames(F) <- marker_genes
colnames(F) <- paste0("k",1:7)
factors <- paste0("k",c(4,7,3,5,2,6))
annotation_heatmap(F[,factors],select_features = "all",verbose = FALSE)
```

Let's now try with $K = 13$ factors:

```{r flashier-nmf-k-13, cache=TRUE, warning=FALSE}
fl_nmf <- flash(shifted_log_counts,S = s1,ebnm_fn = ebnm_point_exponential,
                var_type = 2,greedy_Kmax = 13,backfit = FALSE,
				nullcheck = FALSE,verbose = 2)
fl_nmf <- flash_backfit(fl_nmf,extrapolate = FALSE,maxiter = 40,verbose = 3)
fl_nmf <- flash_backfit(fl_nmf,extrapolate = TRUE,maxiter = 80,verbose = 3)
```

ADD TEXT HERE.

```{r structure-plot-flashier-nmf-k-13, fig.height=2.5, fig.width=6.5, results="hide", message=FALSE}
L <- ldf(fl_nmf,type = "i")$L
colnames(L) <- paste0("k",1:13)
structure_plot(L[,-1],grouping = clusters,gap = 10)
```
