---
title: Comparing the Poisson log1p NMF and topic model on the Stancill 2021 LSA data
author: Eric Weine
output: workflowr::wflow_html
---

This is our initial NMF analysis of ther Stancill 2021 LSA data set.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the packages used in this analysis:

```{r load-packages, message=FALSE}
library(Matrix)
library(readr)
library(dplyr)
library(fastTopics)
library(log1pNMF)
library(ggplot2)
library(cowplot)
```

Load the data set:

```{r load-data}
load("../data/pancreas_cytokine_lsa.Rdata")
barcodes   <- as.data.frame(barcodes)
clusters   <- factor(barcodes$celltype,
                     c("Acinar","Ductal","Endothelial/Mesnchymal","Macrophage",
			  	       "Alpha","Beta","Delta","Gamma"))
conditions <- factor(barcodes$condition,
                     c("Untreated","IL-1B","IFNg","IL-1B_IFNg"))
```

Fit the log1p models:

```{r, eval=FALSE}
cc_vec <- c(1)

K_vec <- c(12, 15)

init_method_vec <- c("rank1")

for (init_method in init_method_vec) {

  for (K in K_vec) {

    for (cc in cc_vec) {

      set.seed(1)
      fit <- fit_poisson_log1p_nmf(
        Y = counts,
        K = K,
        cc = cc,
        init_method = init_method,
        loglik = "exact",
        control = list(maxiter = 250)
      )

      readr::write_rds(
        fit,
        glue::glue("results/lsa_pancreas_cytokine_log1p_c{cc}_{init_method}_init_K{K}.rds")
      )

    }

  }

}
```

Fit the topic models:

```{r, eval=FALSE}
set.seed(1)
tm_k12 <- fit_poisson_nmf(
  X = as(counts, "CsparseMatrix"),
  k = 12,
  control = list(nc = 7)
)

set.seed(1)
tm_k15 <- fit_poisson_nmf(
  X = as(counts, "CsparseMatrix"),
  k = 15,
  control = list(nc = 7)
)
```


## Topic model, K = 12

Here, I attempt to separate out fitted topics based on their
association with either celltype or treatment.

```{r load-topics-k12}
tm_k12 <- read_rds("../output/panc_cyto_lsa_res/tm_k12.rds")
```

The topic model appears to have more factors that are specific to
celltype, and fewer that seem to capture the treatment effects.

```{r structure-plot-topic-model-k12, fig.height=5.5, fig.width=6.5, results="hide", message=FALSE}
set.seed(1)
celltype_topics <- c(2:3,5:11)
other_topics <- c(1,4,12)
topic_colors <- fastTopics:::kelly()[c(2:12,14)]
i <- c(sample(which(clusters == "Beta"),800),
       which(clusters != "Beta"))
L <- poisson2multinom(tm_k12)$L
p1 <- structure_plot(L[i,],grouping = clusters[i],gap = 15,n = Inf,
                     topics = celltype_topics,colors = topic_colors)	
p2 <- structure_plot(L[i,],grouping = clusters[i],gap = 15,n = Inf,
                     topics = other_topics,colors = topic_colors)
p3 <- structure_plot(L[i,],grouping = conditions[i],gap = 20,n = Inf,
                     topics = other_topics,colors = topic_colors)
plot_grid(p1,p2,p3,nrow = 3,ncol = 1,rel_heights = c(3.3,3,2))
```

Topic 1 appears to be non-IL-1B specific, whereas topic 4 appears to
be a general cytokine effect (but perhaps moreso capturing the effects
of IL-1B).

## Poisson log1p NMF, K = 12

Let's now look at the results from the Poisson log1p NMF, with $K =
12$.

```{r load-log1p-nmf-k12}
log1p_k12 <- read_rds(file.path("../output/panc_cyto_lsa_res",
               "lsa_pancreas_cytokine_log1p_c1_rank1_init_K12.rds"))
```

Here are the factors that appeared most correlated with cell type.

```{r structure-plot-log1p-nmf-k12, fig.height=6, fig.width=6.5, results="hide", message=FALSE}
scale_cols <- function (A, b)
  t(t(A) * b)
set.seed(1)
celltype_topics = c(1,2,3,5,6,7,8,9)
topic_colors <- fastTopics:::kelly()[c(2:12,14)]
L <- log1p_k12$LL
d <- apply(L,2,max)
L <- scale_cols(L,1/d)
p1 <- structure_plot(L[i,],grouping = clusters[i],gap = 15,n = Inf,
                     topics = celltype_topics,colors = topic_colors) +
  labs(y = "membership",fill = "")					 
p2 <- structure_plot(L[i,],grouping = clusters[i],gap = 15,n = Inf,
                     topics = c(4,10,11,12),colors = topic_colors) +
  labs(y = "membership",fill = "")
p3 <- structure_plot(L[i,],grouping = conditions[i],gap = 30,n = Inf,
                     topics = c(4,10,11,12),colors = topic_colors) +
  labs(y = "membership",fill = "")					 
plot_grid(p1,p2,p3,nrow = 3,ncol = 1)
```

NOTE: UPDATE THIS TEXT. The pockets of purple and yellow in the beta
cell cluster may represent cells that were misclassified (though it is
very hard to tell). Here, it appears that the pink factor is specific
to conditions that are not treated with IL-1B, where the blue factor
is specific to conditions treated with IL-1B. The orange factor
appears specific to conditions treated with IFNg. The condition with
both treatments has both the orange and the blue factors.

## K = 15

```{r, eval=FALSE}
tm_fit15 <- readr::read_rds("../data/panc_cyto_lsa_res/tm_k15.rds")
```

```{r, eval=FALSE}
structure_plot(tm_fit15, grouping = barcodes$celltype, gap = 20)

# Treatment specific factors:
# k2
# k8
# k9
# k10
# k12
# k14
# k15

# celltype specific factors:
# k1
# k3
# k4
# k5
# k6
# k7
# k11
# k13
```

```{r, eval=FALSE}
structure_plot(
  tm_fit15, grouping = barcodes$condition, gap = 20,
  topics = paste0("k", c(2, 8, 9, 10, 12, 14, 15))
  )
```

Here, it appears that the pink factor (k10) and grey (k8) is mostly
specific to the untreated condition, the dark orange (k14) is specific
to IL-1B, the maroon (k15) is represented in all treated conditions.

```{r, eval=FALSE}
log1p_fit15 <- readr::read_rds("../data/panc_cyto_lsa_res/lsa_pancreas_cytokine_log1p_c1_rank1_init_K15.rds")
```

```{r, eval=FALSE}
normalized_structure_plot(log1p_fit15, grouping = barcodes$celltype, gap = 20, n = Inf)
```

