---
title: Annotation of pancreas factors
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we re-examine some of the [matrix factorization results from
the pancreas CEL-seq2 data](pancreas_another_look.html),
with the goal of understanding how best to *annotate* the pancreas
factors. As we will see, there isn't a single "one-size-fits-all"
strategy that works best, so it is suggested that several annotation
strategies be explored.

The plotting functions used in this analysis are from
[fastTopics][fastTopics].

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

First, load the packages needed for this analysis.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(flashier)
library(fastTopics)
library(ggplot2)
library(cowplot)
```

Set the seed for reproducibility.

```{r set-seed}
set.seed(1)
```

## Structure plot

[fastTopics]: https://github.com/stephenslab/fastTopics/
