---
title: Comparing the log1p and topic models on Stancill 2021 LSA
author: Eric Weine
output: workflowr::wflow_html
---

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

```{r, include=FALSE}
library(Matrix)
library(fastTopics)
library(dplyr)
library(log1pNMF)
```

## K = 12

Here, I attempt to separate out fitted topics based on their association with either celltype or treatment. I'll start with the log1p model with $K = 12$.

```{r, include=FALSE}
load("../data/panc_cyto_lsa.Rdata")
```

```{r, include=FALSE}
fit_log1p_12 <- readr::read_rds("../data/panc_cyto_lsa_res/lsa_pancreas_cytokine_log1p_c1_rank1_init_K12.rds")
```

Here are the factors that appeared most correlated with cell type.

```{r, include=FALSE}
p <- normalized_structure_plot(
  fit_log1p_12,
  grouping = barcodes$celltype,
  topics = c(
    paste0("k", "_", c(1, 2, 3, 5, 6, 7, 8, 9))
  ),
  gap = 20
)
```

```{r}
p
```

The pockets of purple and yellow in the beta cell cluster may represent cells that were misclassified (though it is very hard to tell).

And here are those that are most correlated with treatment condition:

```{r, include=FALSE}
p <- normalized_structure_plot(
  fit_log1p_12,
  grouping = barcodes$condition,
  topics = c(
    paste0("k", "_", c(4, 10, 11, 12))
  ),
  gap = 20
)
```

```{r}
p
```

Here, it appears that the pink factor is specific to conditions that are not treated with IL-1B, where the blue factor is specific to conditions treated with IL-1B. The orange factor appears specific to conditions treated with IFNg. The condition with both treatments has both the orange and the blue factors.

Next, we look at the topic model:

```{r, include=FALSE}
tm_fit <- readr::read_rds("../data/panc_cyto_lsa_res/tm_k12.rds")
```

The topic model appears to have more factors that are specific to celltype, and fewer that seem to capture the treatment effects.

Below are the celltype specific factors:

```{r, include=FALSE}
p <- structure_plot(
  tm_fit,
  n = Inf,
  grouping = barcodes$celltype,
  gap = 20,
  topics = paste0("k", c(2, 3, 5, 6, 7, 8, 9, 10, 11, 12))
)
```

```{r}
p
```

Only factor 4 (and possibly 1?) appeared to be correlated with treatment:

```{r, include=FALSE}
p <- structure_plot(
  tm_fit,
  n = Inf,
  grouping = barcodes$condition,
  gap = 40,
  topics = paste0("k", c(1, 4))
)
```

```{r}
p
```

Factor 1 appears to be non-IL-1B specific, where factor 4 appears to be a general cytokine effect (but perhaps moreso capturing the effects of IL-1B).

## K = 15

```{r, include=FALSE}
tm_fit15 <- readr::read_rds("../data/panc_cyto_lsa_res/tm_k15.rds")
```

```{r, eval=FALSE}
structure_plot(tm_fit15, grouping = barcodes$celltype, gap = 20)

# Treatment specific factors:
# k2
# k8
# k9
# k10
# k12
# k14
# k15

# celltype specific factors:
# k1
# k3
# k4
# k5
# k6
# k7
# k11
# k13
```
```{r, eval=FALSE}
structure_plot(
  tm_fit15, grouping = barcodes$condition, gap = 20,
  topics = paste0("k", c(2, 8, 9, 10, 12, 14, 15))
  )

```
Here, it appears that the pink factor (k10) and grey (k8) is mostly specific to the untreated condition, the dark orange (k14) is specific to IL-1B, the maroon (k15) is represented in all treated conditions.


```{r, eval=FALSE}
log1p_fit15 <- readr::read_rds("../data/panc_cyto_lsa_res/lsa_pancreas_cytokine_log1p_c1_rank1_init_K15.rds")
```

```{r, eval=FALSE}
normalized_structure_plot(log1p_fit15, grouping = barcodes$celltype, gap = 20, n = Inf)
```

