---
title: NMF analysis of the "pancreas cytokine" data set (untreated mouse only)
author: Peter Carbonetto
output: workflowr::wflow_html
---

Here we perform a NMF analyses of the
["pancreas cytokine" data set](pancreas_cytokine.html), focussing on
the scRNA-seq data from untreated mouse only.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load packages used to process the data, perform the analyses, and
create the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(fastTopics)
library(NNLM)
library(flashier)
library(singlecelljamboreeR)
library(ggplot2)
library(cowplot)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the prepared data set:

```{r load-data}
load("../data/pancreas_cytokine.RData")
```

Here we will analyze the cells from the untreated mouse only:

```{r filter-cells-1}
i       <- which(samples$mouse == "S1")
samples <- samples[i,]
counts  <- counts[i,]
```

Remove three cells that appear to be outliers (one of them appears to
be an acinar cell based on Eric's analysis):

```{r filter-cells-2}
outliers <- c("TTTGTTGTCGTTAGTG-1","TTTGTTGGTAGAGCTG-1","CCCAACTCACTCATAG-1")
i        <- which(!is.element(samples$barcode,outliers))
samples  <- samples[i,]
counts   <- counts[i,]
```

Remove genes that are expressed in fewer than 5 cells:

```{r filter-genes}
j      <- which(colSums(counts > 0) > 4)
genes  <- genes[j,]
counts <- counts[,j]
```

This is the dimension of the data set we will analyze:

```{r dim-counts}
dim(counts)
```

For the Gaussian-based analyses (later), we will need the shifted log
counts:

```{r shifted-log-counts}
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
shifted_log_counts <- log1p(counts/(a*s))
rownames(shifted_log_counts) <- NULL
```

## Topic model (fastTopics)

Fit a topic model to the counts (with $K = 13$ topics):

```{r fit-topic-model, cache=TRUE}
set.seed(1)
tm <- fit_poisson_nmf(counts,k = 13,init.method = "random",method = "em",
                      numiter = 40,verbose = "none",
					  control = list(numiter = 4,nc = 8,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 40,
                      control = list(numiter = 4,nc = 8,extrapolate = TRUE),
					  verbose = "none")
```

Structure plots comparing the topics to the clusters (some of which
are inferred from the topics):

```{r structure-plot-topic-model, fig.height=4.25, fig.width=6.5, results="hide", message=FALSE}
set.seed(1)
celltype_topics <- paste0("k",c(2,3,5,7:9,13))
other_topics <- paste0("k",c(4,1,6,10:12))
L <- poisson2multinom(tm)$L
clusters <- as.character(samples$cluster)
clusters[clusters == "islet"]                  <- "beta"
clusters[clusters == "beta" & L[,"k3"] > 0.25] <- "alpha"
clusters[clusters == "beta" & L[,"k8"] > 0.25] <- "delta+epsilon"
clusters[clusters == "beta" & L[,"k9"] > 0.25] <- "gamma"
clusters <- factor(clusters,c("beta","alpha","delta+epsilon","gamma","duct",
                              "endothelial-mesenchymal","macrophage"))
i <- c(sample(which(clusters == "beta"),400),
       which(clusters != "beta"))
p1 <- structure_plot(L[i,],grouping = clusters[i],topics = celltype_topics,
                     gap = 10,n = Inf) +
  labs(fill = "")
p2 <- structure_plot(L[i,],grouping = clusters[i],topics = other_topics,
                     gap = 10,n = Inf) +
  labs(fill = "")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```

Based on the estimated $\mathbf{F}$, we have the following potential
interpretation of these topics:

+ topic 2: $\beta$ cells (*Ins1*, *Ins2*, *Mafa*)
+ topic 3: $\alpha$ cells (*Gcg*, *Mafb*)
+ topic 5: duct cells (*Krt19*)
+ topic 6: replicate effect
+ topic 7: macrophages (*Ccr5*)
+ topic 8: $\delta$ cells (*Sst*)
+ topic 9: $\gamma$ cells (*Ppy*)
+ topic 12: cell cycle
+ topic 13: endothelial (*Pecam1*) and mesenchymal cells (*Col1a1*).

Topic 10 is clearly capturing a technical difference in the two
replicates:

```{r k10-vs-replicate, fig.height=1.75, fig.width=2}
pdat <- cbind(samples,L)
ggplot(pdat,aes(x = replicate,y = k10)) +
  geom_boxplot() +
  theme_cowplot(font_size = 10)
```

## EBNMF (flashier)

Next we will fit an NMF to the *shifted log counts* using flashier,
with $K = 13$. Since the greedy initialization does not seem to work
well in this example, I'll use a different initialization strategy:
obtain a "good" initialization using the NNLM package, then use this
initialization to fit a NMF using flashier. This approach is
implemented in the following function:

```{r flashier-nmf-function}
flashier_nmf <- function (X, k = 3, n.threads = 1) {
  n <- nrow(X)
  m <- ncol(X)
  Y <- as.matrix(X)
  nmf0 <- nnmf(Y,k = 1,loss = "mse",method = "scd",max.iter = 10,
               verbose = 2,n.threads = n.threads)
  W0 <- nmf0$W
  H0 <- nmf0$H
  W0 <- cbind(W0,matrix(runif(n*(k-1)),n,k-1))
  H0 <- rbind(H0,matrix(runif(m*(k-1)),k-1,m))
  nmf <- nnmf(Y,k,init = list(W = W0,H = H0),loss = "mse",method = "scd",
              max.iter = 10,verbose = 2,n.threads = n.threads)
  x  <- rpois(1e7,1/n)
  s1 <- sd(log(x + 1))
  out <- flash_init(X,var_type = 2,S = s1)
  out <- flash_factors_init(out,list(nmf$W,t(nmf$H)),ebnm_point_exponential)
  out <- flash_backfit(out,extrapolate = FALSE,maxiter = 100,verbose = 2)
  return(flash_backfit(out,extrapolate = TRUE,maxiter = 100,verbose = 2))
}
```

Now fit an NMF to the shifted log counts, with $K = 13$:

```{r flashier-nmf, cache=TRUE, warning=FALSE, cache=TRUE}
set.seed(1)
fl_nmf <- flashier_nmf(shifted_log_counts,k = 13,n.threads = 8)
```

Structure plot comparing the factors to the clusters:

```{r structure-plot-flashier-nmf, fig.height=4.5, fig.width=6.5, results="hide", message=FALSE}
set.seed(1)
celltype_factors <- paste0("k",c(1:5,9:11,13))
other_factors <- paste0("k",c(6:8,12))
L <- ldf(fl_nmf,type = "i")$L
colnames(L) <- paste0("k",1:13)
i <- c(sample(which(clusters == "beta"),400),
       which(clusters != "beta"))
p1 <- structure_plot(L[i,],grouping = clusters[i],topics = celltype_factors,
                    gap = 10,n = Inf) +
  labs(y = "membership",fill = "")
p2 <- structure_plot(L[i,],grouping = clusters[i],topics = other_factors,
                    gap = 10,n = Inf) +
  labs(y = "membership",fill = "")
print(plot_grid(p1,p2,nrow = 2,ncol = 1))
```

Based on the estimated $\mathbf{F}$, we have the following potential
interpretation of these topics:

+ factor 1: $\beta$ cells (*Ins1*, *Ins2*, *Mafa*)
+ factors 2, 9, 11: endothelial cells (*Pecam1*, *Esam*)
+ factor 3: duct cells (*Krt19*)
+ factor 4: macrophages (*Ccr5*)
+ factor 5: $\alpha$ (*Gcg*, *Mafb*) and $\gamma$ cells (*Ppy*)
+ factor 6: replicate effect
+ factors 7, 12: cell cycle
+ factor 10: mesenchymal cells (*Col1a1*)
+ factor 13: $\delta$ cells (*Sst*)

Factor 6 almost perfectly captures the technical difference in the two
replicates:

```{r k6-vs-replicate, fig.height=2, fig.width=2}
pdat <- cbind(samples,L)
r <- cor(as.numeric(samples$replicate),L[,"k6"])
ggplot(pdat,aes(x = replicate,y = k6)) +
  geom_boxplot() +
  ggtitle(paste("cor =",round(r,digits = 3))) +
  theme_cowplot(font_size = 12)
```

Factors 7 and 12 seem to capture [cell cycle][reactome-cell-cycle]:

```{r cell-cycle}
cell_cycle <- 
c("Aaas", "Abl1", "Abraxas1", "Acd", "Actr1a", "Adrm1", "Ahctf1",
"Ajuba", "Akap9", "Akt1", "Akt2", "Akt3", "Alms1", "Anapc1",
"Anapc10", "Anapc11", "Anapc15", "Anapc16", "Anapc2", "Anapc4",
"Anapc5", "Anapc7", "Ankle2", "Ankrd28", "Arpp19", "Atm", "Atrip",
"Atrx", "Aurka", "Aurkb", "B9d2", "Babam1", "Babam2", "Banf1",
"Bard1", "Blm", "Blzf1", "Bora", "Brca1", "Brcc3", "Brip1", "Bub1b",
"Bub3", "Cables1", "Cc2d1b", "Ccna2", "Ccnb1", "Ccnb2", "Ccnd1",
"Ccnd2", "Ccnd3", "Ccne1", "Ccne2", "Ccnh", "Ccp110", "Cdc14a",
"Cdc16", "Cdc20", "Cdc23", "Cdc25a", "Cdc25b", "Cdc26", "Cdc27",
"Cdc45", "Cdc6", "Cdc7", "Cdca5", "Cdca8", "Cdk1", "Cdk11b",
"Cdk2", "Cdk4", "Cdk5rap2", "Cdk6", "Cdk7", "Cdkn1a", "Cdkn1b",
"Cdkn1c", "Cdkn2b", "Cdt1", "Cenpa", "Cenpc1", "Cenpe", "Cenpf",
"Cenph", "Cenpi", "Cenpj", "Cenpk", "Cenpl", "Cenpm", "Cenpn",
"Cenpo", "Cenpp", "Cenpq", "Cenps", "Cenpt", "Cenpu", "Cenpw",
"Cenpx", "Cep131", "Cep135", "Cep152", "Cep164", "Cep192", "Cep250",
"Cep290", "Cep41", "Cep57", "Cep63", "Cep70", "Cep72", "Cep76",
"Cep78", "Cetn2", "Chek1", "Chek2", "Chmp2a", "Chmp2b", "Chmp3",
"Chmp4b", "Chmp4c", "Chmp6", "Chmp7", "Chtf18", "Chtf8", "Ckap5",
"Cks1b", "Clasp1", "Clasp2", "Clip1", "Clspn", "Cnep1r1", "Cop1",
"Csnk1d", "Csnk1e", "Csnk2a1", "Csnk2a2", "Csnk2b", "Ctc1", "Ctdnep1",
"Cul1", "Daxx", "Dbf4", "Dctn1", "Dctn2", "Dctn3", "Dkc1", "Dmc1",
"Dna2", "Dscc1", "Dsn1", "Dync1h1", "Dync1i1", "Dync1i2", "Dync1li1",
"Dync1li2", "Dynll1", "Dynll2", "Dyrk1a", "E2f1", "E2f2", "E2f3",
"E2f4", "E2f5", "Emd", "Eml4", "Ercc6l", "Esco1", "Esco2", "Espl1",
"Exo1", "Fbxl18", "Fbxl7", "Fbxo5", "Fbxw11", "Fen1", "Fignl1",
"Fkbpl", "Foxm1", "Fzr1", "Gar1", "Gins1", "Gins2", "Gins3",
"Gins4", "Gmnn", "Golga2", "Gorasp1", "Gorasp2", "Gtse1", "H3f3a",
"H3f3b", "Haus1", "Haus2", "Haus3", "Haus4", "Haus5", "Haus6",
"Haus7", "Haus8", "Hdac1", "Hdac8", "Herc2", "Hjurp", "Hmmr",
"Hsp90aa1", "Hsp90ab1", "Hus1", "Incenp", "Ist1", "Itgb3bp",
"Jak2", "Kat5", "Kif18a", "Kif20a", "Kif23", "Kif2a", "Kif2c",
"Kntc1", "Kpnb1", "Lbr", "Lcmt1", "Lig1", "Lin37", "Lin52", "Lin54",
"Lin9", "Lmna", "Lmnb1", "Lpin2", "Lpin3", "Lyn", "Mad1l1", "Mad2l1",
"Mapk1", "Mapk3", "Mapre1", "Mastl", "Mau2", "Mcm10", "Mcm2",
"Mcm3", "Mcm4", "Mcm5", "Mcm6", "Mcm7", "Mcm8", "Mcph1", "Mdc1",
"Mdm2", "Mdm4", "Mis12", "Mis18a", "Mis18bp1", "Mnat1", "Mre11a",
"Mzt1", "Mzt2", "Nbn", "Ncapd2", "Ncapd3", "Ncapg", "Ncapg2",
"Ncaph", "Ncaph2", "Ndc1", "Ndc80", "Nde1", "Ndel1", "Nedd1",
"Nek2", "Nek9", "Nhp2", "Ninl", "Nipbl", "Nme7", "Nop10", "Npm1",
"Nsd2", "Nsl1", "Nudc", "Nuf2", "Numa1", "Nup107", "Nup133",
"Nup153", "Nup155", "Nup160", "Nup188", "Nup205", "Nup210", "Nup214",
"Nup35", "Nup37", "Nup43", "Nup50", "Nup54", "Nup62", "Nup85",
"Nup88", "Nup93", "Nup98", "Odf2", "Ofd1", "Oip5", "Optn", "Orc2",
"Orc3", "Orc4", "Orc5", "Orc6", "Pafah1b1", "Pcm1", "Pcna", "Pds5a",
"Pds5b", "Phf20", "Phlda1", "Pias4", "Pkmyt1", "Plk1", "Plk4",
"Pmf1", "Pola1", "Pola2", "Pold1", "Pold2", "Pold3", "Pold4",
"Pole", "Pole2", "Pole3", "Pole4", "Pom121", "Pot1a", "Ppme1",
"Ppp1cb", "Ppp1cc", "Ppp1r12a", "Ppp1r12b", "Ppp2ca", "Ppp2cb",
"Ppp2r1a", "Ppp2r1b", "Ppp2r2a", "Ppp2r3d", "Ppp2r5a", "Ppp2r5b",
"Ppp2r5c", "Ppp2r5d", "Ppp2r5e", "Ppp6c", "Ppp6r3", "Prim1",
"Prim2", "Prkaca", "Prkcb", "Psma1", "Psma2", "Psma3", "Psma4",
"Psma5", "Psma6", "Psma7", "Psmb1", "Psmb2", "Psmb3", "Psmb4",
"Psmb5", "Psmb6", "Psmb7", "Psmc1", "Psmc2", "Psmc3", "Psmc4",
"Psmc5", "Psmc6", "Psmd1", "Psmd11", "Psmd12", "Psmd13", "Psmd14",
"Psmd2", "Psmd3", "Psmd6", "Psmd7", "Psmd8", "Pttg1", "Rab1a",
"Rab1b", "Rab2a", "Rab8a", "Rad1", "Rad17", "Rad21", "Rad50",
"Rad51", "Rad9a", "Rad9b", "Rae1", "Ran", "Ranbp2", "Rangap1",
"Rb1", "Rbbp4", "Rbbp7", "Rbbp8", "Rbl1", "Rbl2", "Rbm39", "Rbx1",
"Rcc1", "Rcc2", "Rfc1", "Rfc2", "Rfc3", "Rfc4", "Rfc5", "Rhno1",
"Rmi1", "Rmi2", "Rnf168", "Rnf8", "Rpa1", "Rpa2", "Rpa3", "Rsf1",
"Rtel1", "Ruvbl1", "Sdccag8", "Sec13", "Seh1l", "Set", "Sfi1",
"Sfn", "Sgo1", "Sgo2a", "Shq1", "Sirt2", "Ska2", "Skp2", "Smarca5",
"Smc1a", "Smc2", "Smc3", "Smc4", "Spast", "Spc24", "Spc25", "Spdl1",
"Src", "Ssna1", "Stag1", "Stag2", "Stn1", "Sumo1", "Taok1", "Ten1",
"Terf1", "Terf2", "Terf2ip", "Tert", "Tfdp1", "Top3a", "Topbp1",
"Tpr", "Tpx2", "Trp53", "Trp53bp1", "Tuba1a", "Tuba1b", "Tuba1c",
"Tuba4a", "Tuba8", "Tubb2a", "Tubb2b", "Tubb3", "Tubb4a", "Tubb4b",
"Tubb5", "Tubb6", "Tubg1", "Tubg2", "Tubgcp2", "Tubgcp3", "Tubgcp4",
"Tubgcp5", "Tubgcp6", "Uba52", "Ubb", "Ubc", "Ube2c", "Ube2d1",
"Ube2e1", "Ube2i", "Ube2n", "Ube2s", "Ube2v2", "Uimc1", "Vps4a",
"Vrk1", "Vrk2", "Wapl", "Wee1", "Wrap53", "Wrn", "Xpo1", "Ywhab",
"Ywhae", "Ywhag", "Ywhah", "Ywhaq", "Ywhaz", "Zfp385a", "Zw10",
"Zwilch", "Zwint")
F <- ldf(fl_nmf,type = "i")$F
rownames(F) <- genes$symbol
colnames(F) <- paste0("k",1:13)
sort(colSums(F[cell_cycle,]))
```

## Closer examination of the topics and factors

First, we confirm that many of the "cluster-like" topics and factors
are very similar to each other. Note that the topics are shown on the
(base-10) log scale.

```{r topics-vs-factors, fig.height=4, fig.width=6, warning=FALSE}
topic_vs_factor_scatterplot <- function (factor, topic, title = "") {
  pdat <- data.frame(factor = factor,
                     topic  = log10(topic + 1e-5))
  r <- cor(pdat$factor,pdat$topic)
  return(ggplot(pdat,aes_string(x = "factor",y = "topic")) +
         geom_point(color = "black") +
		 ggtitle(title) +
         theme_cowplot(font_size = 10) +
		 theme(plot.title = element_text(face = "plain",size = 10)))
}
F1 <- ldf(fl_nmf,type = "i")$F
F2 <- poisson2multinom(tm)$F
rownames(F1) <- genes$symbol
colnames(F1) <- paste0("k",1:13)
rownames(F2) <- genes$symbol
p1 <- topic_vs_factor_scatterplot(F1[,"k1"],F2[,"k2"],"beta cells")
p2 <- topic_vs_factor_scatterplot(F1[,"k5"],F2[,"k3"],"alpha cells")
p3 <- topic_vs_factor_scatterplot(F1[,"k13"],F2[,"k8"],"delta cells")
p4 <- topic_vs_factor_scatterplot(F1[,"k3"],F2[,"k5"],"ductal cells")
p5 <- topic_vs_factor_scatterplot(F1[,"k4"],F2[,"k7"],"macrophages")
plot_grid(p1,p2,p3,p4,p5,nrow = 2,ncol = 3)
```

These next scatterplots show that the topics remain very much
"cluster-like" in the sense that much of the top genes by expression
are shared across the different topics. Here to illustrate I am
comparing topic 8 (the "delta cells" topic) to other topics. Notice
that even fairly unrelated cell types such as ductal cells (k5) and
macrophages (k7) still have a lot of shared expression.

```{r gene-matrix-scatterplots-topics, fig.height=4, fig.width=6}
gene_matrix_scatterplot <- function (gene_matrix, k1, k2) {
  gene_matrix <- as.data.frame(gene_matrix)
  cor(gene_matrix)
  return(ggplot(gene_matrix,aes_string(x = k1,y = k2)) +
         geom_point(color = "black") +
         geom_abline(intercept = 0,slope = 1,color = "orangered",
	                 linetype = "dashed") +
		 ggtitle(paste("cor =",
		               round(cor(gene_matrix[,k1],
		                         gene_matrix[,k2]),digits = 3))) +
         theme_cowplot(font_size = 10) +
		 theme(plot.title = element_text(face = "plain",size = 10)))
}
F2 <- log10(F2 + 1e-5)
p1 <- gene_matrix_scatterplot(F2,"k8","k2")
p2 <- gene_matrix_scatterplot(F2,"k8","k3")
p4 <- gene_matrix_scatterplot(F2,"k8","k5")
p3 <- gene_matrix_scatterplot(F2,"k8","k7")
p5 <- gene_matrix_scatterplot(F2,"k8","k9")
p6 <- gene_matrix_scatterplot(F2,"k8","k13")
plot_grid(p1,p2,p3,p4,p5,p6,nrow = 2,ncol = 3)
```

Less so for the factors, particularly for factors that are less
related, such as the delta cells topic (k13) vs. ductal cells (k3)
or delta cells (k13) vs. macrophages (k4):

```{r gene-matrix-scaterplots-factors, fig.height=4, fig.width=4}
p1 <- gene_matrix_scatterplot(F1,"k13","k1")
p2 <- gene_matrix_scatterplot(F1,"k13","k3")
p3 <- gene_matrix_scatterplot(F1,"k13","k5")
p4 <- gene_matrix_scatterplot(F1,"k13","k4")
plot_grid(p1,p2,p3,p4,nrow = 2,ncol = 2)
```

The cell cycle factors (k7, k12) are other examples where the factors
are very independent from the others, e.g.,

```{r gene-matrix-scatterplots-k7-factor, fig.height=4, fig.width=6}
p1 <- gene_matrix_scatterplot(F1,"k7","k1")
p2 <- gene_matrix_scatterplot(F1,"k7","k3")
p3 <- gene_matrix_scatterplot(F1,"k7","k4")
p4 <- gene_matrix_scatterplot(F1,"k7","k5")
p5 <- gene_matrix_scatterplot(F1,"k7","k13")
plot_grid(p1,p2,p3,p4,p5,nrow = 2,ncol = 3)
```

TO DO:

+ Based on these results, identify "top genes" for the topics and
  factors, either as "distinctive" genes (using de_analysis for topic
  model and with a custom function from the singlecelljamboreeR
  package for the factors), or as simply the genes with the largest
  entries in the F matrix. I first need to determine which of the two
  strategies to use for each topic and factor. Then create two
  annotation plots for genes based on these strategies: one for the
  topic model and one for the EBNMF.

+ Also perform GSEA using these results.

[reactome-cell-cycle]: https://www.gsea-msigdb.org/gsea/msigdb/mouse/geneset/REACTOME_CELL_CYCLE.html
