---
title: "pancreas_celseq2_gbcd"
author: "Matthew Stephens"
date: "2025-07-02"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(gbcd)
library(flashier)
library(fastTopics)
library(Matrix)
library(ggplot2)
library(cowplot)
library(ggrepel)
library(RcppML)
```

## Introduction

I want to look at the gbcd fit to the Pancreas celseq2 data.
The 

First I load the data and select the CEL-seq2 data, which should select 2,285 cells.
Then I remove genes that are expressed in fewer than 10 cells, and compute the shifted log counts.

```{r}
load("../data/pancreas.RData")
set.seed(1)

# Select the CEL-seq2 data (Muraro et al, 2016).
# This should select 2,285 cells.
i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i,]
counts      <- counts[i,]

# Remove genes that are expressed in fewer than 10 cells.
x      <- colSums(counts > 0)
j      <- which(x > 9)
counts <- counts[,j]

# Compute the shifted log counts.
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
Y <- MatrixExtra::mapSparse(counts/(a*s),log1p)

#randomly divide rows of Y into 2
subset = sample(1:nrow(Y), nrow(Y)/2)

#dat <- Matrix::tcrossprod(Y)/ncol(Y) # cell by cells covariance matrix
```

Now load the gbcd fits created by `fit_pancreas_celseq2_gbcd.R`. This loads 3 fits: fit.gbcd is the whole dataset and fit.gbcd.1 and fit.gbcd2 are fits to two random halves of the data (subset and its complement).
```{r}
load("../output/pancreas_celseq2_gbcd.RData")
```

Plot the structure plot for the whole fit:
```{r structure-plots-celseq2-flashier, fig.height=4, fig.width=8, message=FALSE, results="hide"}

sample_info <- transform(sample_info,celltype = factor(celltype))
celltype <- sample_info$celltype
celltype <-
 factor(celltype,
        c("acinar","ductal","activated_stellate","quiescent_stellate",
          "endothelial","macrophage","mast","schwann","alpha","beta",
          "delta","gamma","epsilon"))

L <- fit.gbcd$L

p1 <- structure_plot(L,grouping = celltype,
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")
plot(p1)
```

Then two other structure plots:
```{r}
L.1 <- fit.gbcd.1$L

p1 <- structure_plot(L.1,grouping = celltype[subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")
plot(p1)
```


```{r}
L.2 <- fit.gbcd.2$L

p2 <- structure_plot(L.2,grouping = celltype[-subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")
plot(p2)
```

Compute the cross correlations of the factors from the two independent fits. (Note that the rows of xcor correspond to the first fit and the columns to the second fit.) 

A histogram shows that xcors exceeding 0.5 are 
"above the noise".
```{r}
xcor = cor(fit.gbcd.1$F$lfc,fit.gbcd.2$F$lfc)
hist(xcor)
```

For each row of xcor (GEPs in first fit) find the names of the columns (GEPs in second fit) whose correlation exceed 0.5.
The matches are those GEPs that have exactly one such match (a few have 2 other GEPs that are correlated >0.5 suggesting they are somewhat split).
```{r}
above_noise = apply(xcor,1,function(x) names(x)[x > 0.5])
matches = above_noise[lapply(above_noise,length)==1]
```
  
  
## Structure Plots with matching GEPs
  
```{r}
p1.match <- structure_plot(L.1[,names(matches)],grouping = celltype[subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")

p2.match <- structure_plot(L.2[,unlist(matches)],grouping = celltype[-subset],
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")

plot_grid(p1.match,p2.match,ncol=1)
```

  
## Volcano plots of matching GEPs  
Volcano plot code.
```{r}
volcano_plot=function(res.gbcd,GEP){
  pdat <- data.frame(gene = rownames(res.gbcd$F$lfc), 
                   lfc = res.gbcd$F$lfc[, GEP], 
                   z = abs(res.gbcd$F$z_score[, GEP]), 
                   lfsr = res.gbcd$F$lfsr[, GEP],
                   stringsAsFactors = FALSE)
  pdat <- transform(pdat, lfsr = cut(lfsr, c(-1, 0.001, 0.01, 0.05, Inf)))
  rows  <- with(pdat, which(!(abs(lfc) > quantile(abs(lfc), 0.998) | (z > 10))))
  pdat[rows, "gene"] <- ""
  p = ggplot(pdat, aes(x = lfc, y = z, color = lfsr, label = gene)) + geom_point() + 
    geom_text_repel(color = "black", size = 2.5, segment.color = "black",
                  segment.size = 0.25, min.segment.length = 0,
                  max.overlaps = Inf, na.rm = TRUE) +
  scale_color_manual(values = c("coral", "orange", "gold", "deepskyblue")) +
  labs(x = "log-fold change", y = "|posterior z-score|") + 
  guides(colour = guide_legend(override.aes = list(size = 2))) + 
  theme(plot.title      = element_text(hjust = 0.5,size = 12),
        axis.text       = element_text(size = 10),
        axis.title      = element_text(size = 10),
        legend.title    = element_text(size = 12),
        legend.text     = element_text(size = 10),
        legend.position = "bottom") +
  ggtitle(paste0(GEP))
  return(p)
}
```

For each match plot the GEPs of both fits next to one another:
```{r}
for(GEP.1 in names(matches)){
  print(GEP.1)
  GEP.2 <- matches[[GEP.1]]
  p.1 <- volcano_plot(fit.gbcd.1, GEP.1)
  p.2 <- volcano_plot(fit.gbcd.2, GEP.2)
  p = plot_grid(p.1, p.2,nrow=1)
  plot(p)
}
```

I was interested to notice that GEP20, which corresponds to Gamma (PP) cells, has MAFB being downregulated. Further inspection shows that indeed MAFB is low expressed in Gamma cells in these data. That is not something I have found noted in the literature.

Eyeballing the plots, most seem to match quite well; exceptions are maybe GEP15-GEP21 which don't match so well; also some differences between GEP22-GEP22? But overall subsetting to the highly correlated GEPs seems to work well at creating a pretty consistent structure plot.


## Try regular NMF

Since I could not find a convenient implementation of symmetric NMF in R I used regular nmf on the Gram matrix (note that because it is symmetric the result tends to be close to symmetric, but it would be nice to have a version that properly respects symmetry.)

```{r}
dat <- Matrix::tcrossprod(Y)/ncol(Y) # cell by cells covariance matrix 
temp = nmf(as.matrix(dat),10)
p1 <- structure_plot(temp@w,grouping = celltype,
                     gap = 20,perplexity = 70,n = Inf) +
  labs(y = "membership",fill = "factor",color = "factor")
plot(p1)
```


### Try Symmetric NMF

Here I try Annie Xie's R code for symNMF based on the method of Vandaele et al. This is work in progress.

```{r}
# this is an implementation of the coordinate descent symNMF method from A. Vandaele, N. Gillis, Q. Lei, K. Zhong, I. Dhillon
# given an n-by-n symmetric nonnegative matrix A and a factorization rank r, it computes a n-by-r nonnegative matrix H
# which solves the problem min_{H >0} \|A - HH'\|_F^2

CoDesymNMF_init <- function(A, r, init_type){
  n <- ncol(A)
  if(init_type == 'zeros'){
    H <- matrix(rep(0,n*r), ncol = r)
  }else{
    H <- matrix(runif(n*r), ncol = r)
  }
}

CoDesymNMF_fit <- function(A, H, maxiter = 100, rand_permutation = FALSE){
  n <- nrow(H)
  r <- ncol(H)
  
  # scaling (if H is not the zero matrix)
  if (max(H) > 0){
    HHt <- tcrossprod(H)
    beta <- sum(A*HHt)/sum(HHt^2) #scaling factor
    H <- sqrt(beta)*H
  }
  
  iter <- 1
  x <- rep(0,4)
  # main loop
  while(iter <= maxiter){
    # randomly permute columns if rand_permutation is TRUE
    if (rand_permutation == TRUE){
      col_permutation <- sample(r, size = r, replace = FALSE)
      H <- H[,col_permutation]
    }
    
    # calculate residual
    R <- A - tcrossprod(H)
    for (k in 1:r){
      R <- R + tcrossprod(H[,k])
      diag_R <- diag(R)
      HtH <- crossprod(H[,k]) #scalar, equals sum(H[,k]^2)
      for (i in 1:n){
        # coefficients a and b of x^3 + ax + b
        HtH <- HtH - H[i,k]^2 #scalar
        a <- HtH - diag_R[i] #scalar
        b <- -(t(H[,k]) %*% R[,i] - H[i,k]*diag_R[i]) #scalar
        
        Delta <- 4*(a^3) + 27*(b^2) # note: delta can be negative!
        d <- 0.5 * (-b + sqrt(as.complex(Delta/27))) #might be complex
        
        if (Delta <= 0) {
          r3 <- 2*(abs(d)^(1/3))
          th3 <- atan2(Im(d),Re(d))/3
          x[2] <- r3*cos(th3)
          x[3] <- r3*cos(th3 + (2*pi/3))
          x[4] <- r3*cos(th3 + (4*pi/3))
          x <- x[x >= 0] 
          ind <- which.min((x^4)/4 + c(a)*(x^2)/2 + c(b)*x)
          H[i,k] <- x[ind]
          HtH <- HtH + H[i,k]^2
        }else{
          d <- Re(d) # d is just a real number in this case
          z <- sign(d)*(abs(d))^(1/3)
          val <- z - (a/(3*z))
          if(((val^4)/4 + a*(val^2/2) + b*val < 0) & (val >= 0)){
            HtH <- HtH + val^2
            H[i,k] <- val
          }else{
            H[i,k] <- 0
          }
        }
      } # end of 1:n for loop
      R <- R - tcrossprod(H[,k])
    } # end of 1:r for loop 
    iter <- iter + 1
  } # end of maxiter while loop
  # return output
  return(list(H=H, obj_func = sum((A - tcrossprod(H))^2)))
}

CoDesymNMF <- function(A, r, init_type = 'zeros', maxiter = 100, rand_permutation = FALSE){
  H_init <- CoDesymNMF_init(A, r, init_type)
  CoDesymNMF_results <- CoDesymNMF_fit(A, H_init, maxiter, rand_permutation)
  return(CoDesymNMF_results)
}
```

