---
title: "pancreas_celseq2_ica"
author: "Matthew Stephens"
date: "2025-10-27"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r}
library(fastICA)
library("Matrix")
library(ggplot2)
```

## Introduction

I wanted to try fastICA on the pancreas data. I also experiment with a "warm start" 
using gradient steps that *minimize* log cosh, because the minima of log-cosh tend to correspond to sparse binary (0,1) groups rather than sign (-1,1) groups that can combine clusters.

```{r}
#fits ica to the pancreas celseq2 data (and 2 random subsets)
load("../data/pancreas.RData")
set.seed(1)

# Select the CEL-seq2 data (Muraro et al, 2016).
# This should select 2,285 cells.
i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i,]
counts      <- counts[i,]

# Remove genes that are expressed in fewer than 10 cells.
x      <- colSums(counts > 0)
j      <- which(x > 9)
counts <- counts[,j]

# Compute the shifted log counts.
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
Y <- MatrixExtra::mapSparse(counts/(a*s),log1p)

#randomly divide rows of Y into 2
subset = sample(1:nrow(Y), nrow(Y)/2)

```


Helper functions: 
```{r}
# matrix version of r1 fastica
# U is (n.comp+1) x n (whitened data plus intercept).  
# W is (n.comp+1) x n_starts (one weight vector per column).
# P = t(U) %*% W is n x n_starts (source estimates).
fastica_update = function(U, W) {
  P  <- t(U) %*% W          # n x n_starts: source estimates
  G  <- tanh(P)
  G2 <- 1 - G^2
  W  <- U %*% G - sweep(W, 2, colSums(G2), "*")
  # Add epsilon (1e-15) to prevent 0/0
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}

  
# X is n x p; returns whitened data (centers columns of X, whiten to n.comp dimensions)
# returned U is n.comp x n 
preprocess = function(X, n.comp = 10) {
  X <- scale(X, scale = FALSE)
  sqrt(nrow(X)) * t(svd(X)$u[, 1:n.comp])
}

gradient_minica_update = function(U, W, lr = 0.1) {
  # n_samples = ncol(U); n_features = nrow(U); n_starts = ncol(W)
  
  P <- t(U) %*% W          # n_samples x n_starts: source estimates
  G <- tanh(P)             # First derivative of log-cosh
  
  # Calculate the gradient 
  # We divide by ncol(U) to average over samples, keeping the learning rate 
  # stable regardless of your dataset size.
  grad <- (U %*% G) / ncol(U)
  
  # Gradient descent step (minimize log-cosh)
  # Note: To MAXIMIZE log-cosh (standard for super-Gaussian sources), use + instead of -
  W <- W - lr * grad
  
  # Normalize to project back to the unit sphere (norm = 1)
  # Add epsilon (1e-15) to prevent 0/0
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}

# this is the projected version; i have not tested since the unprojected version seems to work fine
gradient_minica_update_projected = function(U, W, lr = 0.1) {
  # 1. Compute the Euclidean gradient
  P <- t(U) %*% W
  G <- tanh(P)
  grad <- (U %*% G) / ncol(U)
  
  # 2. Project gradient onto the tangent space of the sphere
  # Calculate the dot product (w^T g) for each column
  w_T_grad <- colSums(W * grad) 
  
  # Subtract the radial component: g_proj = g - (w^T g) * w
  proj_grad <- grad - sweep(W, 2, w_T_grad, "*")
  
  # 3. Take the gradient step using the projected gradient
  W <- W - lr * proj_grad
  
  # 4. Retract back to the unit sphere
  sweep(W, 2, sqrt(colSums(W^2)) + 1e-15, "/")
}

# A greedy function to prune correlated columns of a matrix; written by Gemini
# Install if necessary: install.packages("caret")
library(caret)

fast_prune_caret <- function(L, tau = 0.8) {
  # Compute correlation matrix once
  cor_mat <- abs(cor(L))
  
  # findCorrelation returns the indices to REMOVE
  # Setting exact = FALSE uses a faster heuristic for large matrices
  drop_indices <- findCorrelation(cor_mat, cutoff = tau, exact = FALSE)
  
  # Handle the case where no rows exceed the threshold
  if (length(drop_indices) > 0) {
    kept_indices <- setdiff(1:ncol(L), drop_indices)
    pruned_matrix <- L[,-drop_indices,drop = FALSE]
  } else {
    kept_indices <- 1:ncol(L)
    pruned_matrix <- L
  }
  
  list(pruned_matrix = pruned_matrix, kept_indices = kept_indices)
}


prune_and_count_cluster <- function(L, tau = 0.9) {
  # Drop constant columns before computing correlation (zero variance -> NaN in cor())
  L <- L[, apply(L, 2, sd) > 1e-10, drop = FALSE]
  # 1. Convert correlation to distance (1 - absolute correlation)
  dist_mat <- as.dist(1 - abs(cor(L)))
  
  # 2. Hierarchical clustering 
  # 'complete' linkage ensures no two rows in a cluster are further apart than the threshold
  hc <- hclust(dist_mat, method = "complete")
  
  # 3. Cut the dendrogram to form clusters (distance of 1 - tau corresponds to correlation of tau)
  clusters <- cutree(hc, h = 1 - tau)
  
  # 4. Select a representative from each cluster 
  # match() quickly grabs the first index of each unique cluster ID
  kept_indices <- match(unique(clusters), clusters)
  
  # 5. Map the sizes to the exact order of kept_indices
  # Extract the cluster ID for each kept row, then use it to index the table
  cluster_sizes_table = table(clusters)
  kept_cluster_ids <- clusters[kept_indices]
  cluster_sizes <- as.integer(cluster_sizes_table[as.character(kept_cluster_ids)])
  
  list(
    pruned_matrix = L[,kept_indices , drop = FALSE],
    cluster_sizes = cluster_sizes,
    kept_indices = kept_indices,
    cluster_assignments = clusters
  )
}

celltype_palette <- c(
  "#E41A1C", "#377EB8", "#4DAF4A", "#984EA3", "#FF7F00",
  "#A65628", "#F781BF", "#1B9E77", "#D95F02", "#7570B3",
  "#E7298A", "#66A61E", "#E6AB02", "#A6761D", "#666666"
)

lhat_ggplot <- function(Lhat.pc, idx, title) {
  if (sum(idx) == 0) return(invisible(NULL))
  Lhat.prune <- Lhat.pc$pruned_matrix
  Lhat_sub   <- Lhat.prune[, idx, drop = FALSE]
  obj_sub    <- colMeans(log(cosh(Lhat_sub)))
  o2         <- order(obj_sub)
  cell_order <- order(sample_info$celltype)
  n_cells    <- nrow(Lhat_sub)
  n_comp     <- sum(idx)

  comp_labels <- make.unique(paste0("n:", Lhat.pc$cluster_sizes[idx], " obj:", round(obj_sub, 3)))
  df <- data.frame(
    rank      = rep(seq_len(n_cells), n_comp),
    celltype  = rep(sample_info$celltype[cell_order], n_comp),
    loading   = as.vector(Lhat_sub[cell_order, ]),
    component = factor(rep(comp_labels, each = n_cells), levels = comp_labels[o2])
  )
  ggplot(df, aes(x = rank, y = loading, color = celltype)) +
    geom_point(size = 0.5, alpha = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
    facet_wrap(~ component, ncol = 5, scales = "free_y") +
    scale_color_manual(values = celltype_palette) +
    labs(x = NULL, y = "Loading", color = "Cell type", title = title) +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          strip.text = element_text(size = 7, margin = margin(2, 0, 2, 0)),
          strip.background = element_rect(fill = "grey90", color = NA),
          legend.position = "bottom") +
    guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
}

plot_Lhat_maxima <- function(Lhat.pc, obj_threshold = 0.42) {
  obj <- colMeans(log(cosh(Lhat.pc$pruned_matrix)))
  lhat_ggplot(Lhat.pc, obj > obj_threshold, paste0("Maxima (obj > ", obj_threshold, ")"))
}

plot_Lhat_minima <- function(Lhat.pc, obj_threshold = 0.42) {
  obj <- colMeans(log(cosh(Lhat.pc$pruned_matrix)))
  lhat_ggplot(Lhat.pc, obj <= obj_threshold, paste0("Minima (obj <= ", obj_threshold, ")"))
}

# General version for unlabeled/unordered loading matrices
plot_loadings_matrix <- function(Lhat, label_prefix = "ct:") {
  obj <- colMeans(log(cosh(Lhat)))
  cell_order <- order(sample_info$celltype)
  n_cells <- nrow(Lhat)
  n_comp <- ncol(Lhat)

  comp_labels <- make.unique(paste0(label_prefix, seq_len(n_comp), " obj:", round(obj, 3)))
  df <- data.frame(
    rank      = rep(seq_len(n_cells), n_comp),
    celltype  = rep(sample_info$celltype[cell_order], n_comp),
    loading   = as.vector(Lhat[cell_order, ]),
    component = factor(rep(comp_labels, each = n_cells), levels = comp_labels)
  )

  ggplot(df, aes(x = rank, y = loading, color = celltype)) +
    geom_point(size = 0.5, alpha = 0.7) +
    geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.3) +
    facet_wrap(~ component, ncol = 5, scales = "free_y") +
    scale_color_manual(values = celltype_palette) +
    labs(x = NULL, y = "Loading", color = "Cell type") +
    theme_bw(base_size = 10) +
    theme(axis.text.x = element_blank(), axis.ticks.x = element_blank(),
          strip.text = element_text(size = 7, margin = margin(2, 0, 2, 0)),
          strip.background = element_rect(fill = "grey90", color = NA),
          legend.position = "bottom") +
    guides(color = guide_legend(override.aes = list(size = 3, alpha = 1)))
}
```

## Eigenvalue plots

First plot the eigenvalues of Y to get some idea how many components to whiten to:
```{r}
Y     <- scale(Y, scale = FALSE)  # center columns; also densifies for SVD
Y.svd <- svd(Y)
df_scree <- data.frame(k = 2:1000, d = Y.svd$d[2:1000])

ggplot(df_scree, aes(k, d)) +
  geom_point(size = 0.5) +
  labs(x = "Component", y = "Singular value") +
  theme_bw()

ggplot(df_scree[df_scree$k <= 200, ], aes(k, d)) +
  geom_point(size = 0.5) +
  geom_vline(xintercept = 30, color = "red", linetype = "dashed") +
  labs(x = "Component", y = "Singular value") +
  theme_bw()
```


Whiten data:
```{r}
n.comp = 25 # I found using slightly fewer that 30 PCs produced maybe better results
U     <- sqrt(nrow(Y)) * t(Y.svd$u[, 1:n.comp])
U_aug <- rbind(rep(1, ncol(U)), U)
```


## Initialize from random normal starts

I run fastICA (rank 1) from 1000 random normal starts and then cluster the results (using hierarchical clustering). The plot shows the number of starts that gave each result and the objective value obtained, with panels ordered by increasing objective value. (I also tried minimizing from binary starts but this did not change the results much; if the binary starts were very unbalanced then they tended to converge more often to the intercept.) 

```{r}
n_starts = 1000
n_iter = 50  #you can get away with fewer
set.seed(1)
W <- matrix(rnorm(nrow(U_aug) * n_starts), nrow(U_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- fastica_update(U_aug, W)

Lhat <- t(U_aug) %*% W  # n x n_starts

Lhat.pc = prune_and_count_cluster(Lhat)
Lhat.prune = Lhat.pc$pruned_matrix

table(sample_info$celltype)

plot_Lhat_maxima(Lhat.pc)
plot_Lhat_minima(Lhat.pc)
```

## Minimizing log cosh: random init

Here I try use the gradient warmstart to minimize log cosh from 1000 different starting points. In this case the warm start (50 iterations) is enough to make all runs converge to local minima. This basically finds all the minima that the original did, plus one more (objective 0.246 which corresponds to delta cells)

```{r}
n_starts = 1000
n_iter = 50  
set.seed(1)
W <- matrix(rnorm(nrow(U_aug) * n_starts), nrow(U_aug), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- gradient_minica_update(U_aug, W)
for (i in seq_len(n_iter))
  W <- fastica_update(U_aug, W)

Lhat <- t(U_aug) %*% W  # n x n_starts

Lhat.pc = prune_and_count_cluster(Lhat)
Lhat.prune = Lhat.pc$pruned_matrix

plot_Lhat_minima(Lhat.pc)
```


## Initialize from known celltypes: binary

Here I make a binary (0/1) matrix with one column for each cell type, and initialize the ica from that, using the warm start to minimize (which actually does not make much difference in this case; not shown). I wanted to see if there were local minima, corresponding to specific cell types, that were missed in the above random starts. 
The only result found here that is missing from the random starts is the component corresponding to beta cells (ct4, objective 0.303).


```{r}
X_bin = model.matrix(~ sample_info$celltype - 1)

W <- U_aug %*% X_bin
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- gradient_minica_update(U_aug, W)
for (i in seq_len(n_iter))
  W <- fastica_update(U_aug, W)

Lhat_ct <- t(U_aug) %*% W  # n x n_starts
plot_loadings_matrix(Lhat_ct[, 1:13])
```

## Initialize from known celltypes: sign

Here I make a sign (-1/1) matrix with one column for each cell type, and initialize the ica from that, just to see which of these splits are stable. The biggest difference from the 0/1 initialization is the alpha cells (ct3): here the split remains stable, but the minimization moved away from the split. One possibility is that the fact that these cells are quite common is penalizing them in the minimization which prefers sparser groups. It may be interesting to run the minimization with fewer alpha cells. A couple of other cell types (acinar, ct1; ductal, ct6) remain much more binary in this case than in the corresponding 0/1 minima. It may be interesting to see how these behave under ELBO maximization with unbalanced binary priors.

```{r}
X_sign = 2*X_bin-1 

W <- U_aug %*% X_sign
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- fastica_update(U_aug, W)

Lhat_ct <- t(U_aug) %*% W  # n x n_starts
plot_loadings_matrix(Lhat_ct[, 1:13])
```



## Initialize by splitting the original results

Many of the original solutions split about -1,1, and are close to a maximum of log cosh (around 0.43). From simulation results we know that some of these may be combining groups.
Here I try initializing at the 0,a version of these solutions, again using warmstart to minimize. It finds most of the ones found from random starts (all except the split that corresponds to gamma cells, obj 0.216), but no additional solutions.

```{r}
X = (sign(Lhat.prune)+1)/2
W = U_aug %*% cbind(X,1-X)

W <- sweep(W, 2, sqrt(colSums(W^2))+1e-15, "/")
for (i in seq_len(n_iter))
  W <- gradient_minica_update(U_aug, W)
for (i in seq_len(n_iter))
  W <- fastica_update(U_aug, W)

Lhat2 <- t(U_aug) %*% W  # n x n_starts

obj2 = colMeans(log(cosh(Lhat2)))
Lhat2.pc = prune_and_count_cluster(Lhat2)

plot_Lhat_minima(Lhat2.pc)

```


## Run with fewer alpha cells

Here I try running with fewer alpha cells. Indeed it now finds a local minima that corresponds to the split of alpha cells vs others. This seems to point to a limitation of the fastICA with local minima. (However, it is possible that these larger groups may emerge when we have corrected for the many smaller groups?)
```{r}
set.seed(1)

# Thin alpha cells: keep all non-alpha cells + 200 random alpha cells
alpha_idx     <- which(sample_info$celltype == "alpha")
keep_alpha    <- sample(alpha_idx, 200)
thin_idx      <- sort(c(which(sample_info$celltype != "alpha"), keep_alpha))

sample_info_thin <- sample_info[thin_idx, ]
counts_thin      <- counts[thin_idx, ]

# Recompute size-factor normalisation and log1p transform for thinned data
s_thin  <- rowSums(counts_thin) / mean(rowSums(counts_thin))
Y_thin  <- MatrixExtra::mapSparse(counts_thin / (a * s_thin), log1p)

# Centre and SVD
Y_thin       <- scale(Y_thin, scale = FALSE)
Y_thin.svd   <- svd(Y_thin)
U_thin       <- sqrt(nrow(Y_thin)) * t(Y_thin.svd$u[, 1:n.comp])
U_aug_thin   <- rbind(rep(1, ncol(U_thin)), U_thin)

# 1000 random starts
W <- matrix(rnorm(nrow(U_aug_thin) * n_starts), nrow(U_aug_thin), n_starts)
W <- sweep(W, 2, sqrt(colSums(W^2)), "/")
for (i in seq_len(n_iter))
  W <- gradient_minica_update(U_aug_thin, W)
for (i in seq_len(n_iter))
  W <- fastica_update(U_aug_thin, W)

Lhat_thin    <- t(U_aug_thin) %*% W
Lhat_thin.pc <- prune_and_count_cluster(Lhat_thin)

# Plot using thinned sample_info (swap temporarily so plotting functions work)
sample_info_save <- sample_info
sample_info      <- sample_info_thin
plot_Lhat_minima(Lhat_thin.pc)
sample_info      <- sample_info_save
```


This was old code, running it to cluster genes and looking how consistent the programs are from the two runs. While there is some consistency, there do not seem to be as many consistent programs as with semi-nmf (eg not as many with abs correlation >0.5). Note that, unlike with semi-nmf, the programs will not necessarily have a consistent sign.
```{r}
#fit.ica.k40 = fastICA(t(Y), n.comp=40)
#fit.ica.k40.1 = fastICA(t(Y[subset,]), n.comp = 40)
#fit.ica.k40.2 = fastICA(t(Y[-subset,]), n.comp = 40)

#session_info <- sessionInfo()

#save(list = c("fit.snmf.k40","fit.snmf.k40.1","fit.snmf.k40.2","session_info"),
 #    file = "../output/pancreas_celseq2_snmf_k40.RData")


# cormat <- cor(fit.ica.k40.1$S,fit.ica.k40.2$S)
# apply(abs(cormat),1, max)
# hist(cormat,nclass=100)
# image(abs(cormat)>0.5)
# assignment_problem <- RcppHungarian::HungarianSolver(-1*abs(cormat))
# pairings <- assignment_problem$pairs
# image(abs(cormat)[pairings[,1], pairings[,2]])

```


